---
title: "Executive Summary: Motor Insurance Portfolio Analysis"
author: "Huang Kuan Yu"
date: "May 2026"

format:
  revealjs:
    theme: simple
    slide-number: true
    width: 1600
    height: 900
    margin: 0.04
    center: false
    preview-links: auto
    css: slides.css
    auto-stretch: false

execute:
  echo: false
  warning: false
  message: false
  freeze: true
---

```{r}
#| include: false

library(tidyverse)
library(readxl)
library(scales)
library(gganimate)
library(showtext)

font_add_google("Cabin Sketch", "Cabin Sketch")
font_add_google("DM Sans", "DM Sans")
showtext_auto()

primary_color <- "#A8D8EA"
secondary_color <- "#FFF2A8"
third_color <- "#B8E0D2"
fourth_color <- "#D7C6FF"
fifth_color <- "#FFD1DC"

text_dark <- "#4a4a4a"

theme_va <- function() {
  theme_minimal() +
    theme(
      text = element_text(family = "DM Sans", color = text_dark),
      plot.title = element_text(family = "Cabin Sketch", size = 28, face = "bold", color = text_dark, hjust = 0.5),
      axis.title = element_text(family = "DM Sans", size = 14, color = text_dark),
      axis.text = element_text(family = "DM Sans", size = 12, color = text_dark),
      legend.title = element_text(family = "DM Sans", size = 12, color = text_dark),
      legend.text = element_text(family = "DM Sans", size = 11, color = text_dark),
      panel.grid.major = element_line(color = "#eeeeee", linewidth = .4),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

motor_raw <- read_delim(
  "data/Dataset of motor insurance portfolio.csv",
  delim = ";",
  show_col_types = FALSE
)

motor_clean <- motor_raw %>%
  mutate(
    year = factor(year),
    policy_type = factor(policy_type),
    policy_status = factor(policy_status),
    fuel_type = factor(fuel_type),
    vehicle_brand = factor(vehicle_brand),
    driver_age = as.numeric(driver_age),
    total_claims = as.numeric(total_claims),
    total_incurred = as.numeric(total_incurred),
    total_exposure = as.numeric(total_exposure),
    total_premium = as.numeric(total_premium)
  )

policy_year_metrics <- motor_clean %>%
  group_by(year, policy_type) %>%
  summarise(
    total_claims = sum(total_claims, na.rm = TRUE),
    total_incurred = sum(total_incurred, na.rm = TRUE),
    total_exposure = sum(total_exposure, na.rm = TRUE),
    total_premium = sum(total_premium, na.rm = TRUE),
    policy_count = n(),
    .groups = "drop"
  )

status_year_metrics <- motor_clean %>%
  mutate(
    status_group = if_else(policy_status == "A", "Active", "Not Active")
  ) %>%
  group_by(year, status_group) %>%
  summarise(
    total_claims = sum(total_claims, na.rm = TRUE),
    total_incurred = sum(total_incurred, na.rm = TRUE),
    total_exposure = sum(total_exposure, na.rm = TRUE),
    total_premium = sum(total_premium, na.rm = TRUE),
    policy_count = n(),
    .groups = "drop"
  )
```

## Contents {.text-slide}

1.  Study Objectives\
2.  Data Preparation\
3.  Portfolio Concentration\
4.  Driver Demographics\
5.  Retention and Policy Activity Trends\
6.  Claim Frequency Across Policy Types\
7.  Underwriting Profitability by Policy Type\
8.  Activity Status and Risk Performance\
9.  Claim Severity and Financial Impact\
10. Driver Age and Risk Profiles\
11. Key Findings & Future Research

::: notes
This presentation summarises findings from an exploratory analysis of a motor insurance portfolio. The goal is to understand portfolio composition, retention patterns, claims behaviour and underwriting performance across different policy groups. Throughout the presentation, I highlight factors associated with higher risk, weaker profitability and potential concentration concerns. The findings are exploratory and aim to support future statistical analysis or predictive modelling.
:::

::: footnote-box
Full report:

https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html
:::

------------------------------------------------------------------------

## Study Objectives: Understanding Portfolio Risk and Profitability {.text-slide}

### Objectives

- Analyse portfolio composition and demographics
- Examine retention behaviour
- Evaluate claims experience and profitability
- Identify factors associated with risk and volatility

::: footnote-box
Further details:

[Overview — Full Report](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html)
:::

::: notes
This analysis investigates relationships between policy characteristics and insurance performance. Specifically, it examines portfolio composition, demographic exposure, retention behaviour, claim frequency and profitability indicators such as loss ratio. The objective is to identify segments associated with elevated risk or weaker underwriting outcomes and to generate insights that may support future pricing strategies, retention efforts or risk management initiatives.
:::

------------------------------------------------------------------------

## Data Preparation Summary {.text-slide}

- Checked missing values and data completeness
- Converted variable types
- Created retention groups
- Aggregated insurance metrics
- Prepared datasets for exploratory analysis

::: footnote-box
Further details:

[Data Wrangling — Full Analysis](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html#data-wrangling)
:::

::: notes
Before analysis, the dataset was cleaned and transformed to ensure variables were correctly typed and suitable for aggregation. Additional metrics and grouped variables, such as active versus inactive policy status, were created to support comparisons across portfolio segments. These preparation steps help ensure subsequent findings are interpretable and consistent.
:::

------------------------------------------------------------------------

## Portfolio Exposure Is Concentrated in Selected Policy Types

```{r}
#| fig-width: 8
#| fig-height: 4
#| out-width: 60%

policy_distribution <- motor_clean %>% mutate(policy_type = recode(policy_type, "TP" = "Basic Third-Party Liability", "TPG" = "Third-Party Liability + Glass", "CC" = "Third-Party + Additional Coverages", "COMP_E" = "Comprehensive with Excess", "COMP_N" = "Comprehensive without Excess")) %>% count(policy_type, sort = TRUE) %>%
  mutate(pct = n / sum(n))

ggplot(policy_distribution, aes(x = reorder(policy_type, n), y = n)) +
  geom_col(fill = primary_color, color = "white", alpha = .9) +
  geom_text(aes(label = paste0(comma(n)," (", percent(pct, accuracy = 0.1),")")), hjust = -0.1, size = 10, color = text_dark, family = "DM Sans") +
  coord_flip() +
  labs(title = "Composition of Policies by Policy Type", x = "Policy Type", y = "Number of Policies") +
  expand_limits(y = max(policy_distribution$n) * 1.1) +
  theme_va()
```

::: footnote-box
- Analyses the distribution of policies across product types to identify portfolio concentration.\
- CC policies dominate exposure, increasing dependence on a few products.

Further details:

[Policy Type Composition — Full Analysis](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html#policy-type-composition)
:::

::: notes
This chart examines how policies are distributed across product categories. The portfolio appears concentrated within CC policies, indicating that overall portfolio performance may depend heavily on a limited number of products. Such concentration can increase exposure to adverse performance if dominant products experience deteriorating claims behaviour or profitability.
:::

------------------------------------------------------------------------

## Portfolio Exposure Is Dominated by Middle-Aged Drivers

```{r}
#| fig-width: 8
#| fig-height: 4
#| out-height: 70vh

ggplot(motor_clean, aes(x = driver_age)) +
  geom_histogram(bins = 30, fill = primary_color, color = "white", alpha = .85) +
  labs(title = "Distribution of Driver Age", x = "Driver Age", y = "Number of Policies") +
  theme_va()
```

::: footnote-box
- Examines the age distribution of insured drivers to understand demographic exposure.\
- Most policies are held by drivers aged 35–60, indicating concentrated exposure.

Further details:

[Driver Age Distribution — Full Analysis](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html#driver-age-distribution)
:::

::: notes
This distribution shows that most insured drivers fall within middle age groups, particularly around ages 35 to 60. Since exposure is concentrated within these demographics, understanding their claims behaviour becomes important when assessing portfolio risk and profitability. Younger and older drivers contribute comparatively less exposure.
:::

------------------------------------------------------------------------

## Inactive Policies Increase Over Time, Suggesting Retention Pressure

```{r}
#| fig-width: 8
#| fig-height: 4


status_year_summary <- motor_clean %>% mutate(status_group = if_else(policy_status == "A", "Active", "Not Active")) %>% count(year, status_group) %>% group_by(year) %>% mutate(pct = n / sum(n)) %>% ungroup()

ggplot(status_year_summary, aes(x = factor(year), y = pct, fill = status_group)) +
geom_col(color = "white", alpha = .9) +
geom_text(aes(label = percent(pct, accuracy = 1)), position = position_stack(vjust = .5), color = text_dark, size = 10, family = "DM Sans") +
scale_fill_manual(values = c("Active" = primary_color, "Not Active" = secondary_color)) +
scale_y_continuous(labels = percent, limits = c(0,1)) +
labs(title = "Policy Status Composition by Year", x = "Year", y = "Percentage of Policies", fill = "Policy Status") +
theme_va()
```

::: footnote-box
- Compares the proportion of active and inactive policies across years to assess retention trends.\
- Inactive policies increase gradually, suggesting potential weakening retention.

Further details:

[Policy Status Composition — Full Analysis](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html#policy-status-composition-across-years)
:::

::: notes
This chart tracks the proportion of active and inactive policies over time as an indicator of retention behaviour. A gradual increase in inactive policies may suggest weakening retention or higher churn within the portfolio. Persistent increases in inactive status could affect long-term premium growth and portfolio stability.
:::

------------------------------------------------------------------------

## Claim Frequency Varies Significantly Across Policy Types

```{r}
#| fig-width: 8
#| fig-height: 4

ggplot(
  policy_year_metrics,
  aes(
    x = factor(year),
    y = total_claims / total_exposure,
    color = policy_type,
    group = policy_type
  )
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 5, stroke = 1) +
  scale_color_manual(values = c(primary_color, secondary_color, third_color, fourth_color, fifth_color)) +
  labs(title = "Claim Frequency by Policy Type", x = "Year", y = "Claim Frequency", color = "Policy Type") +
  theme_va()
```

::: footnote-box
- Compares claim frequency (claims relative to exposure) across policy types over time.\
- COMP_N consistently records higher claim frequency, indicating elevated risk.

Further details:

[Claim Frequency — Full Analysis](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html#claim-frequency-by-policy-type-over-time)
:::

::: notes
Claim frequency measures how often claims occur relative to exposure. Comparing policy types reveals meaningful differences in claims occurrence over time. COMP_N consistently records higher claim frequency, suggesting elevated claims risk compared with other products. Such patterns may indicate differences in underlying risk profiles or policyholder behaviour.
:::

------------------------------------------------------------------------

## Certain Policy Types Show Persistent Underwriting Losses

```{r}
#| fig-width: 8
#| fig-height: 4


ggplot(
  policy_year_metrics,
  aes(
    x=factor(year),
    y=total_incurred/total_premium,
    color=policy_type,
    group=policy_type
  )
)+
  geom_line(linewidth = 1.2) +
  geom_point(size = 5, stroke = 1) +
  scale_color_manual(values = c(primary_color, secondary_color, third_color, fourth_color, fifth_color)) +
  labs(title = "Loss Ratio by Policy Type", x = "Year", y = "Loss Ratio", color = "Policy Type") +
  theme_va()
```

::: footnote-box
- Evaluates underwriting profitability using loss ratio (incurred losses relative to premiums).\
- COMP_N exceeds loss ratio \>1, suggesting persistent underwriting losses.

Further details:

[Loss Ratio — Full Analysis](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html#loss-ratio-by-policy-type-over-time)
:::

::: notes
Loss ratio compares incurred losses against premiums earned and serves as an indicator of underwriting profitability. Values exceeding one imply claims costs surpass premium income. The results suggest that COMP_N exhibits persistently weaker profitability, potentially indicating pricing concerns or higher underlying risk.
:::

------------------------------------------------------------------------

## Inactive Policies Exhibit Higher Risk and Poorer Profitability

::::::: columns
:::: {.column width="50%"}
::: graph-title
Active Policies
:::

```{r}
#| fig-width: 7
#| fig-height: 5
status_metrics_long <- status_year_metrics %>%
  mutate(claim_frequency = total_claims / total_exposure, loss_ratio = total_incurred / total_premium) %>%
  select(year, status_group, claim_frequency, loss_ratio) %>%
  pivot_longer(cols = c(claim_frequency, loss_ratio), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric, "claim_frequency" = "Claim Frequency", "loss_ratio" = "Loss Ratio"))

ggplot(filter(status_metrics_long, status_group == "Active"), aes(x = factor(year), y = value, fill = metric)) +
  geom_col(position = position_dodge(width = .75), width = .65, color = "white", alpha = .9) +
  geom_text(aes(label = round(value, 2)), position = position_dodge(width = .75), vjust = -.3, size = 7, color = text_dark) +
  scale_fill_manual(values = c("Claim Frequency" = primary_color, "Loss Ratio" = secondary_color)) +
  labs(title = "Claim Frequency and Loss Ratio for Active Policies", x = "Year", y = "Metric Value", fill = "Metric") +
  theme_va()
```
::::

:::: {.column width="50%"}
::: graph-title
Not Active Policies
:::

```{r}
#| fig-width: 7
#| fig-height: 5
ggplot(filter(status_metrics_long, status_group == "Not Active"), aes(x = factor(year), y = value, fill = metric)) +
  geom_col(position = position_dodge(width = .75), width = .65, color = "white", alpha = .9) +
  geom_text(aes(label = round(value, 2)), position = position_dodge(width = .75), vjust = -.3, size = 7, color = text_dark) +
  scale_fill_manual(values = c("Claim Frequency" = primary_color, "Loss Ratio" = secondary_color)) +
  labs(title = "Claim Frequency and Loss Ratio for Not Active Policies", x = "Year", y = "Metric Value", fill = "Metric") +
  theme_va()
```
::::
:::::::

::: footnote-box
- Compares claim frequency and loss ratio between active and inactive policy groups.\
- Inactive policies display higher risk indicators and poorer profitability.

Further details: [Activity Status Analysis — Full Analysis](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html#claim-frequency-and-loss-ratio-by-policy-activity-status)
:::

::: notes
This comparison evaluates claim frequency and loss ratio between active and inactive policy groups. Inactive policies consistently display higher values across both metrics, suggesting greater claims occurrence alongside poorer underwriting outcomes. These patterns imply that inactive segments may contribute disproportionately to portfolio risk and losses.
:::

------------------------------------------------------------------------

## Low Claim Frequency Can Still Produce High Loss Severity

```{r}
#| fig-width: 8
#| fig-height: 4

ggplot(
motor_clean,
aes(
x = total_claims + 1,
y = total_incurred + 1
)
)+
geom_jitter(
color = primary_color,
alpha = .15
)+
geom_smooth(
color = secondary_color,
se = FALSE
)+
scale_x_log10()+
scale_y_log10()+
labs(
title = "Claims vs Incurred (Log Scale)",
x = "Total Claims (log scale)", 
y = "Total Incurred (log scale)"
)+
theme_va()
```

::: footnote-box
- Explores the relationship between claim occurrence and incurred losses using log scales.\
- Lower claim counts may still generate disproportionately large losses.

Further details:

[Claims vs Incurred — Full Analysis](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html#claims-vs-incurred)
:::

::: notes
This scatter plot explores the relationship between total claims and incurred losses using logarithmic scales. While more claims generally correspond to higher losses, some observations show substantial losses despite relatively low claim counts. This highlights the importance of considering claim severity alongside claim frequency when assessing risk exposure.
:::

------------------------------------------------------------------------

## Risk Patterns Vary Across Driver Age and Policy Type

```{r}
#| fig-width: 8
#| fig-height: 4

risk_profile <- motor_clean %>%
  mutate(
    claim_frequency = if_else(
      total_exposure > 0,
      total_claims / total_exposure,
      NA_real_
    ),
    loss_ratio = if_else(
      total_premium > 0,
      total_incurred / total_premium,
      NA_real_
    )
  ) %>%
  filter(
    policy_type %in% c("CC", "COMP_N"),
    !is.na(year),
    !is.na(driver_age),
    is.finite(claim_frequency),
    is.finite(loss_ratio)
  ) %>%
  mutate(
    policy_type = recode(
      policy_type,
      "CC" = "Third-Party + Additional Coverages",
      "COMP_N" = "Comprehensive without Excess"
    )
  )

risk_metric <- risk_profile %>%
  group_by(year, policy_type, driver_age) %>%
  summarise(
    loss_ratio = mean(loss_ratio, na.rm = TRUE),
    claim_frequency = mean(claim_frequency, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(loss_ratio, claim_frequency),
    names_to = "metric",
    values_to = "value"
  ) %>%
  group_by(year, policy_type, metric) %>%
  mutate(
    value_scaled = value / max(value, na.rm = TRUE) * 1.2
  ) %>%
  ungroup() %>%
  mutate(
    metric = recode(
      metric,
      "loss_ratio" = "Loss Ratio",
      "claim_frequency" = "Claim Frequency"
    )
  )

ggplot() +
  geom_density(
    data = risk_profile,
    aes(
      x = driver_age,
      y = after_stat(density / max(density) * .8)
    ),
    fill = secondary_color,
    color = secondary_color,
    alpha = .65,
    adjust = 2
  ) +
  geom_segment(
    data = risk_metric,
    aes(
      x = driver_age,
      xend = driver_age,
      y = 0,
      yend = -value_scaled,
      color = metric
    ),
    alpha = .45,
    linewidth = .8
  ) +
  geom_point(
    data = risk_metric,
    aes(
      x = driver_age,
      y = -value_scaled,
      color = metric
    ),
    size = 2.2,
    alpha = .8
  ) +
  scale_color_manual(
    values = c(
      "Claim Frequency" = primary_color,
      "Loss Ratio" = fourth_color
    )
  ) +
  facet_grid(
    metric ~ policy_type
  ) +
  scale_y_continuous(
    limits = c(-1.35, .9),
    breaks = c(-1.2, 0, .8),
    labels = c("Higher Risk", "0", "Age Density")
  ) +
  labs(
    title = "Age-Based Risk Profiles Across Selected Policy Types | Year: {closest_state}",
    subtitle = "Cloud = Driver age density | Rain = Mean risk metric by age",
    x = "Driver Age",
    y = NULL,
    color = "Metric"
  ) +
  transition_states(
    year,
    transition_length = 2,
    state_length = 5
  ) +
  theme_va() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank()
  )
```

::: footnote-box
- Examines how claim frequency and loss ratio vary across driver age and policy type.\
- Risk patterns differ by age group and evolve across years.

Further details:

[Age-Based Risk Profile — Full Analysis](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html#relationship-between-driver-age-claim-frequency-and-loss-ratio)
:::

::: notes
This visual examines how driver age relates to claim frequency and loss ratio across selected policy types over time. The density distribution represents portfolio exposure, while risk metrics indicate relative performance by age group. Patterns suggest that risk is not evenly distributed and may vary according to both demographic characteristics and policy structure.
:::

------------------------------------------------------------------------

## Key Findings and Opportunities for Deeper Investigation {.text-slide}

### Key Findings

- Policy type associated with profitability differences
- COMP_N appears highest risk
- Inactive policies perform worse
- Claim severity contributes to volatility

### Future Work

- Statistical testing
- Predictive modelling
- Churn analysis
- Longitudinal studies

::: footnote-box
Further details:

[Summary & Future Research — Full Analysis](https://isss-608-vaa-black.vercel.app/Take-Home_Ex/Take-Home_Ex01/Take-Home_Ex01.html#summary-and-future-research)
:::

::: notes
Overall, the analysis identifies several recurring themes. Portfolio exposure is concentrated within selected products and demographics, inactive policies exhibit weaker performance, and certain policy types show persistently elevated risk and poorer profitability. These findings remain exploratory, and future work may include statistical testing, predictive modelling and deeper investigation into retention or churn dynamics.
:::
