---
title: "Hands on Exercise 8"
author: "Huang Kuan Yu"
date: "2026-06-10"
date-modified: "last-modified" 
  
format:
  html:
    toc: true
    toc-expand: 1
    toc-title: "Contents"
    fig-width: 12
    fig-height: 8
    code-fold: false
    code-summary: "Show code"
    css: custom.scss
    number-sections: true

execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Chapter 17

### Getting started

Import the necessary packages

```{r}
pacman::p_load(scales, viridis, lubridate, ggthemes,
               gridExtra, readxl, knitr, data.table,
               CGPfunctions, ggHoriPlot, tidyverse)
```

### Import data

```{r}
attacks <- read_csv("data/eventlog.csv")
```

Before data processing, examine the raw data pattern

```{r}
kable(head(attacks))
```

### Data processing

#### Step 1: Deriving weekday and hour of day fields

```{r}
make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(ts, 
                        tz = tz[1], 
                        quiet = TRUE)
  dt <- data.table(source_country = sc,
                   wkday = weekdays(real_times),
                   hour = hour(real_times))
  return(dt)
  }
```

::: code-tip
*Code Details*

- [`ymd_hms()`](https://lubridate.tidyverse.org/reference/ymd_hms.html) and [`hour()`](https://lubridate.tidyverse.org/reference/hour.html) are from [**lubridate**](https://lubridate.tidyverse.org/) package, and

- [`weekdays()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/weekdays) is a **base** R function.
:::

#### Step 2: Deriving the attacks tibble data frame

```{r}
wkday_levels <- c('Saturday', 'Friday', 
                  'Thursday', 'Wednesday', 
                  'Tuesday', 'Monday', 
                  'Sunday')

attacks <- attacks %>%
  group_by(tz) %>%
  do(make_hr_wkday(.$timestamp, 
                   .$source_country, 
                   .$tz)) %>% 
  ungroup() %>% 
  mutate(wkday = factor(
    wkday, levels = wkday_levels),
    hour  = factor(
      hour, levels = 0:23))
```

Then we see the tibble table.

```{r}
kable(head(attacks))
```

### Build the Calendar Heatmaps

```{r}
#| code-fold: true
grouped <- attacks %>% 
  count(wkday, hour) %>% 
  ungroup() %>%
  na.omit()

ggplot(grouped, 
       aes(hour, 
           wkday, 
           fill = n)) + 
geom_tile(color = "white", 
          size = 0.1) + 
theme_tufte(base_family = "Helvetica") + 
coord_equal() +
scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
labs(x = NULL, 
     y = NULL, 
     title = "Attacks by weekday and time of day") +
theme(axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6) )
```

::: obsv-box
*Lesson Learned*

- a tibble data table called *grouped* is derived by aggregating the attack by *wkday* and *hour* fields.

- a new field called *n* is derived by using `group_by()` and `count()` functions.

- `na.omit()` is used to exclude missing value.

- `geom_tile()` is used to plot tiles (grids) at each x and y position. `color` and `size` arguments are used to specify the border color and line size of the tiles.

- [`theme_tufte()`](https://jrnold.github.io/ggthemes/reference/theme_tufte.html) of [**ggthemes**](https://jrnold.github.io/ggthemes/reference/index.html) package is used to remove unnecessary chart junk. To learn which visual components of default ggplot2 have been excluded, you are encouraged to comment out this line to examine the default plot.

- `coord_equal()` is used to ensure the plot will have an aspect ratio of 1:1.

- `scale_fill_gradient()` function is used to creates a two colour gradient (low-high).
:::

### Building Multiple Calendar Heatmaps

#### Step 1: Deriving attack by country object

```{r}
attacks_by_country <- count(
  attacks, source_country) %>%
  mutate(percent = percent(n/sum(n))) %>%
  arrange(desc(n))
```

#### Step 2: Preparing the tidy data frame

```{r}
top4 <- attacks_by_country$source_country[1:4]
top4_attacks <- attacks %>%
  filter(source_country %in% top4) %>%
  count(source_country, wkday, hour) %>%
  ungroup() %>%
  mutate(source_country = factor(
    source_country, levels = top4)) %>%
  na.omit()
```

#### Step 3: Plotting the Multiple Calender Heatmap by using ggplot2 package.

```{r}
#| code-fold: true
ggplot(top4_attacks, 
       aes(hour, 
           wkday, 
           fill = n)) + 
  geom_tile(color = "white", 
          size = 0.1) + 
  theme_tufte(base_family = "Helvetica") + 
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
  facet_wrap(~source_country, ncol = 2) +
  labs(x = NULL, y = NULL, 
     title = "Attacks on top 4 countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 7),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6) )
```

### Plotting Cycle Plot

#### Step 1: Data Import

```{r}
air <- read_excel("data/arrivals_by_air.xlsx")
```

#### Step 2: Deriving month and year fields

```{r}
air$month <- factor(month(air$`Month-Year`), 
                    levels=1:12, 
                    labels=month.abb, 
                    ordered=TRUE) 
air$year <- year(ymd(air$`Month-Year`))
```

#### Step 3: Extracting the target country

```{r}
Vietnam <- air %>% 
  select(`Vietnam`, 
         month, 
         year) %>%
  filter(year >= 2010)
```

#### Step 4: Computing year average arrivals by month

```{r}
hline.data <- Vietnam %>% 
  group_by(month) %>%
  summarise(avgvalue = mean(`Vietnam`))
```

#### Step 5: Plotting the cycle plot

```{r}
#| code-fold: true
ggplot() + 
  geom_line(data=Vietnam,
            aes(x=year, 
                y=`Vietnam`, 
                group=month), 
            colour="black") +
  geom_hline(aes(yintercept=avgvalue), 
             data=hline.data, 
             linetype=6, 
             colour="red", 
             size=0.5) + 
  facet_grid(~month) +
  labs(axis.text.x = element_blank(),
       title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019") +
  xlab("") +
  ylab("No. of Visitors") +
  theme_tufte(base_family = "Helvetica")
```

### Plotting Slopegraph

#### Step 1: Data Import

```{r}
rice <- read_csv("data/rice.csv")
```

#### Step 2: Plotting the slopegraph

```{r}
#| code-fold: true

rice %>% 
  mutate(Year = factor(Year)) %>%
  filter(Year %in% c(1961, 1980)) %>%
  newggslopegraph(Year, Yield, Country,
                Title = "Rice Yield of Top 11 Asian Counties",
                SubTitle = "1961-1980",
                Caption = "Prepared by: Huang Kuan Yu")
```
