---
title: "Hands on Exercise 7"
author: "Huang Kuan Yu"
date: "2026-06-03"
date-modified: "last-modified" 
  
format:
  html:
    toc: true
    toc-expand: 1
    toc-title: "Contents"
    fig-width: 12
    fig-height: 8
    code-fold: true
    code-summary: "Show code"
    css: custom.scss
    number-sections: true

execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Chapter 27: Modelling, Visualising and Analysing Network Data with R

### Installing packages

```{r}
pacman::p_load(igraph, tidygraph, ggraph, 
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts, 
               concaveman, ggforce)
```

### Import and Review the data

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
glimpse(GAStech_edges)
```

### Data Wrangling

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

Review formatted data

```{r}
glimpse(GAStech_edges)
```

### Data Wrangling

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

```{r}
glimpse(GAStech_edges)
```

### Creating network objects using tidygraph

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)
```

```{r}
GAStech_graph
```

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

### Plotting Static Network Graphs with ggraph package

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

#### Fruchterman and Reingold layout

```{r}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

#### Modifying network node

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))

g + theme_graph()
```

#### Modifying edges

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```

#### Creating facet graphs

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')

### uses theme() to change the position of the legend
g + facet_edges(~Weekday)
```

Additional: add frames to the faceted graphs can add the following codes

```{r}
#| eval: false

g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

#### Working with facet_nodes()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

### Network Metrics Analysis

#### Computing centrality indices

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

#### Visualising network metrics

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```

#### Visualising Community

```{r}
g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(
    group_optimal(weights = Weight)),
         betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_mark_hull(
    aes(x, y, 
        group = community, 
        fill = community),  
    alpha = 0.2,  
    expand = unit(0.3, "cm"),  # Expand
    radius = unit(0.3, "cm")  # Smoothness
  ) + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(fill = Department,
                      size = betweenness_measure),
                      color = "black",
                      shape = 21)
  
g + theme_graph()
```

### Building Interactive Network Graph with visNetwork

#### Data Preparation

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

#### Plotting the first interactive network graph

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

#### Working with visual attributes - Nodes

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

#### Working with visual attributes - Edges

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

#### Interactivity

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

## Chapeter 29: Visualising and Analysing Text Data with R: tidytext methods

### Installing packages

```{r}
pacman::p_load(tidytext, widyr, wordcloud, DT,
               ggwordcloud, textplot, lubridate, hms,
               tidyverse, tidygraph, ggraph, igraph)
```

### Importing Multiple Text Files from Multiple Folders

```{r}
news20 <- "data/20news/"
```

```{r}
read_folder <- function(infolder) {
  tibble(file = dir(infolder, 
                    full.names = TRUE)) %>%
    mutate(text = map(file, 
                      read_lines)) %>%
    transmute(id = basename(file), 
              text) %>%
    unnest(text)
}
```

```{r}
raw_text <- tibble(folder = 
                     dir(news20, 
                         full.names = TRUE)) %>%
  mutate(folder_out = map(folder, 
                          read_folder)) %>%
  unnest(cols = c(folder_out)) %>%
  transmute(newsgroup = basename(folder), 
            id, text)
write_rds(raw_text, "data/rds/news20.rds")


```

::: code-tip
- [`read_lines()`](https://readr.tidyverse.org/reference/read_lines.html) of [**readr**](https://readr.tidyverse.org/) package is used to read up to n_max lines from a file.

- [`map()`](https://purrr.tidyverse.org/reference/map.html) of [**purrr**](https://purrr.tidyverse.org/) package is used to transform their input by applying a function to each element of a list and returning an object of the same length as the input.

- [`unnest()`](https://tidyr.tidyverse.org/reference/nest.html) of **dplyr** package is used to flatten a list-column of data frames back out into regular columns.

- [`mutate()`](https://dplyr.tidyverse.org/reference/mutate.html) of **dplyr** is used to add new variables and preserves existing ones;

- [`transmute()`](https://dplyr.tidyverse.org/reference/mutate.html) of **dplyr** is used to add new variables and drops existing ones.

- [`read_rds()`](https://readr.tidyverse.org/reference/read_rds.html) is used to save the extracted and combined data frame as rds file for future use.
:::

### Initial EDA

```{r}
raw_text <- read_rds("data/rds/news20.rds")
raw_text %>%
  group_by(newsgroup) %>%
  summarize(messages = n_distinct(id)) %>%
  ggplot(aes(messages, newsgroup)) +
  geom_col(fill = "lightblue") +
  labs(y = NULL)
```

#### Removing header and automated email signitures

```{r}
cleaned_text <- raw_text %>%
  group_by(newsgroup, id) %>%
  filter(cumsum(text == "") > 0,
         cumsum(str_detect(
           text, "^--")) == 0) %>%
  ungroup()
```

::: code-tip
- [`cumsum()`](https://rdrr.io/r/base/cumsum.html) of base R is used to return a vector whose elements are the cumulative sums of the elements of the argument.

- [`str_detect()`](https://stringr.tidyverse.org/reference/str_detect.html) from **stringr** is used to detect the presence or absence of a pattern in a string.
:::

#### Removing lines with nested text representing quotes from other users

```{r}
cleaned_text <- cleaned_text %>%
  filter(str_detect(text, "^[^>]+[A-Za-z\\d]")
         | text == "",
         !str_detect(text, 
                     "writes(:|\\.\\.\\.)$"),
         !str_detect(text, 
                     "^In article <")
  )
```

::: code-tip
- [`str_detect()`](https://stringr.tidyverse.org/reference/str_detect.html) from **stringr** is used to detect the presence or absence of a pattern in a string.

- [`filter()`](https://dplyr.tidyverse.org/reference/filter.html) of **dplyr** package is used to subset a data frame, retaining all rows that satisfy the specified conditions.
:::

#### Text data processing

```{r}
library(tidyverse) 
library(tidytext)   
usenet_words <- cleaned_text %>%
  unnest_tokens(word, text) %>%
  filter(str_detect(word, "[a-z']$"),
         !word %in% stop_words$word)

usenet_words %>%
  count(word, sort = TRUE)
```

Use below codes to count words within by newsgroup

```{r}
words_by_newsgroup <- usenet_words %>%
  count(newsgroup, word, sort = TRUE) %>%
  ungroup()
```

#### Visualising Words in newsgroups

```{r}
wordcloud(words_by_newsgroup$word,
          words_by_newsgroup$n,
          max.words = 300)
```

```{r}
set.seed(1234)

words_by_newsgroup %>%
  filter(n > 0) %>%
ggplot(aes(label = word,
           size = n)) +
  geom_text_wordcloud() +
  theme_minimal() +
  facet_wrap(~newsgroup)
```

#### Computing tf-idf within newsgroups

```{r}
tf_idf <- words_by_newsgroup %>%
  bind_tf_idf(word, newsgroup, n) %>%
  arrange(desc(tf_idf))
```

#### Visualising tf-idf within newsgroups

```{r}
tf_idf %>%
  filter(str_detect(newsgroup, "^sci\\.")) %>%
  group_by(newsgroup) %>%
  slice_max(tf_idf, 
            n = 12) %>%
  ungroup() %>%
  mutate(word = reorder(word, 
                        tf_idf)) %>%
  ggplot(aes(tf_idf, 
             word, 
             fill = newsgroup)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ newsgroup, 
             scales = "free") +
  labs(x = "tf-idf", 
       y = NULL)
```

#### Counting and correlating pairs of words with the widyr package

```{r}
newsgroup_cors <- words_by_newsgroup %>%
  pairwise_cor(newsgroup, 
               word, 
               n, 
               sort = TRUE)
```

#### Visualising correlation as a network

```{r}
set.seed(2017)

newsgroup_cors %>%
  filter(correlation > .025) %>%
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(alpha = correlation, 
                     width = correlation)) +
  geom_node_point(size = 6, 
                  color = "lightblue") +
  geom_node_text(aes(label = name),
                 color = "red",
                 repel = TRUE) +
  theme_void()
```

#### Bigram

```{r}
bigrams <- cleaned_text %>%
  unnest_tokens(bigram, 
                text, 
                token = "ngrams", 
                n = 2)

### counting bigrams
bigrams_count <- bigrams %>%
  filter(bigram != 'NA') %>%
  count(bigram, sort = TRUE)

### cleaning bigrams
bigrams_separated <- bigrams %>%
  filter(bigram != 'NA') %>%
  separate(bigram, c("word1", "word2"), 
           sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)

### counting again

bigram_counts <- bigrams_filtered %>% 
  count(word1, word2, sort = TRUE)
```

#### Create a network graph from bigram data frame

```{r}
bigram_graph <- bigram_counts %>%
  filter(n > 3) %>%
  graph_from_data_frame()
bigram_graph
```

#### Visualizing a network of bigrams with ggraph

```{r}
set.seed(1234)

a <- grid::arrow(type = "closed", 
                 length = unit(.15,
                               "inches"))

ggraph(bigram_graph, 
       layout = "fr") +
  geom_edge_link(aes(edge_alpha = n), 
                 show.legend = FALSE,
                 arrow = a, 
                 end_cap = circle(.07,
                                  'inches')) +
  geom_node_point(color = "lightblue", 
                  size = 5) +
  geom_node_text(aes(label = name), 
                 vjust = 1, 
                 hjust = 1) +
  theme_void()
```
