---
title: "Hands-on Exercise 3" author: "Huang Kuan Yu" date: "2026-04-27"
date-modified: "last-modified" format: html: toc: true toc-expand: 4 execute:
echo: true eval: true warning: false freeze: true
---

## Chapter 3: Programming Interactive Data Visualisation with R

### Getting started

Hands on exercise 3 requires to install and launch these R packages

- ggiraph for making ‘ggplot’ graphics interactive.

- plotly, R library for plotting interactive statistical graphs.

- DT provides an R interface to the JavaScript library DataTables that create interactive table on html page.

- tidyverse, a family of modern R packages specially designed to support data science, analysis and communication task including creating static statistical graphs.

- patchwork for combining multiple ggplot2 graphs into one figure.

```{r}
pacman::p_load(ggiraph, plotly,
               patchwork, DT, tidyverse)
```

### Import Data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

### Interactive data visualisation - ggiraph method

ggiraph is an htmlwidget and a ggplot2 extension. It allows ggplot graphics to be interactive. Interactive is made with ggplot geometries that can understand three arguments: - Tooltip: a column of data-sets that contain tooltips to be displayed when the mouse is over elements. - Onclick: a column of data-sets that contain a JavaScript function to be executed when elements are clicked. - Data_id: a column of data-sets that contain an id to be associated with elements. If it used within a shiny application, elements associated with an id (data_id) can be selected and manipulated on client and server sides. Refer to this article for more detail explanation.

### Tooltip effect with tooltip aesthetic

The code chunk consists of two parts. First, an ggplot object will be created. Next, girafe() of ggiraph will be used to create an interactive svg object.

::: panel-tabset
### The plot

```{r}
#| echo: false
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

### The code

```{r}
#| eval: false
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```
:::

### Display multiple information on tooltip

The first three lines of codes in the code chunk create a new field called tooltip. At the same time, it populates text in ID and CLASS fields into the newly created field. Next, this newly created field is used as tooltip field as shown in the code of line 7. So when the mouse pointer hover on an data point of interest, the student’s ID and Class will be displayed.

::: panel-tabset
### The Plot

```{r}
#| echo: false
exam_data$tooltip <- c(paste0(
  "Name = ", exam_data$ID,
  "\n Class = ", exam_data$CLASS))
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 8*0.618
)
```

### The Code

```{r}
#| eval: false
exam_data$tooltip <- c(paste0(
  "Name = ", exam_data$ID,
  "\n Class = ", exam_data$CLASS))
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 8*0.618
)
```
:::

### Customising Tooltip style

Code chunk below uses opts_tooltip() of ggiraph to customize tooltip rendering by add css declarations.

::: panel-tabset
###The plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
tooltip_css <- "background-color:white; #<<
font-style:bold; color:black;" #<<
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list(    #<<
    opts_tooltip(    #<<
      css = tooltip_css)) #<<
)
```

### The code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
tooltip_css <- "background-color:white; #<<
font-style:bold; color:black;" #<<
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list(    #<<
    opts_tooltip(    #<<
      css = tooltip_css)) #<<
)
```
:::

Refer to [Customizing girafe objects](https://www.ardata.fr/ggiraph-book/customize.html) to learn more about how to customise ggiraph objects.

### Displaying statistics on tooltip

::: panel-tabset
### The plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths scores:", mean, "+/-", sem)
}
gg_point <- ggplot(data=exam_data,
                   aes(x = RACE),
) +
  stat_summary(aes(y = MATHS,
                   tooltip = after_stat(
                     tooltip(y, ymax))),
    fun.data = "mean_se",
    geom = GeomInteractiveCol,
    fill = "light blue"
  ) +
  stat_summary(aes(y = MATHS),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  )
girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8*0.618)
```

### The code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths scores:", mean, "+/-", sem)
}
gg_point <- ggplot(data=exam_data,
                   aes(x = RACE),
) +
  stat_summary(aes(y = MATHS,
                   tooltip = after_stat(
                     tooltip(y, ymax))),
    fun.data = "mean_se",
    geom = GeomInteractiveCol,
    fill = "light blue"
  ) +
  stat_summary(aes(y = MATHS),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  )
girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8*0.618)
```
:::

### Hover effect with data_id aesthetic

Interactivity: Elements associated with a data_id (i.e CLASS) will be highlighted upon mouse over. Important note: the default value of the hover css is hover_css = “fill:orange;”.

::: panel-tabset
### The plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

### The code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```
:::

### Styling hover effect

Interactivity: Elements associated with a data_id (i.e CLASS) will be highlighted upon mouse over. Important note: different from previous example, in this example the ccs customisation request are encoded directly.

::: panel-tabset
### The plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list(
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```

### The code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list(
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```
:::

### Combining tooltip and hover effect

Interactivity: Elements associated with a data_id (i.e CLASS) will be highlighted upon mouse over. At the same time, the tooltip will show the CLASS.

::: panel-tabset
### The plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = CLASS,
        data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list(
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```

### The code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = CLASS,
        data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list(
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```
:::

### Click effect with onclick

onclick argument of ggiraph provides hotlink interactivity on the web. Interactivity: Web document link with a data object will be displayed on the webbrowser upon mouse click. Note: Click actions must be a string column in the dataset containing valid javascript instructions.

::: panel-tabset
### The plot

```{r}
#| echo: false
#| fig-widith: 12
#| fig-height: 8
exam_data$onclick <- sprintf("window.open(\"%s%s\")",
"https://www.moe.gov.sg/schoolfinder?journey=Primary%20school",
as.character(exam_data$ID))
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(onclick = onclick),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618)
```

### The code

```{r}
#| eval: false
#| fig-widith: 12
#| fig-height: 8
exam_data$onclick <- sprintf("window.open(\"%s%s\")",
"https://www.moe.gov.sg/schoolfinder?journey=Primary%20school",
as.character(exam_data$ID))
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(onclick = onclick),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618)
```
:::

### Coordinated multiple views with ggiraph

Interactivity: When a data point of one of the dotplot is selected, the corresponding data point ID on the second data visualisation will be highlighted too.

In order to build a coordinated multiple views as shown in the example below, the following programming strategy will be used: Appropriate interactive functions of ggiraph will be used to create the multiple views. patchwork function of patchwork package will be used inside girafe function to create the interactive coordinated multiple views.

::: panel-tabset
### The Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
p1 <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim=c(0,100)) +
  scale_y_continuous(NULL,
                     breaks = NULL)
p2 <- ggplot(data=exam_data,
       aes(x = ENGLISH)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim=c(0,100)) +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(code = print(p1 + p2),
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
         )
       )
```

### The code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
p1 <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim=c(0,100)) +
  scale_y_continuous(NULL,
                     breaks = NULL)
p2 <- ggplot(data=exam_data,
       aes(x = ENGLISH)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim=c(0,100)) +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(code = print(p1 + p2),
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
         )
       )
```
:::

### Interactive data visualisation - plotly methods

There are two ways to create interactive graph by using plotly, they are: by using plot_ly(), and by using ggplotly()

### Creating an interactive scatter plot: plot_ly() method vs. ggplotly() method

#### The Plot

::: panel-tabset
### The plot_ly() Method

```{r}
#| echo: false
#| fig-widith: 12
#| fig-height: 8
plot_ly(data = exam_data,
             x = ~MATHS,
             y = ~ENGLISH)
```

### The ggplotly() method

```{r}
#| echo: false
#| fig-widith: 12
#| fig-height: 8
p <- ggplot(data=exam_data,
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
ggplotly(p)
```
:::

#### The Code

::: panel-tabset
### The plot_ly() Method

```{r}
#| eval: false
#| fig-widith: 12
#| fig-height: 8
plot_ly(data = exam_data,
             x = ~MATHS,
             y = ~ENGLISH)
```

### The ggplotly() method

```{r}
#| eval: false
#| fig-widith: 12
#| fig-height: 8
p <- ggplot(data=exam_data,
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
ggplotly(p)
```
:::

### Working with visual variable: plot_ly() method

::: panel-tabset
### The Plot

```{r}
#| echo: false
#| fig-widith: 12
#| fig-height: 8
plot_ly(data = exam_data,
        x = ~ENGLISH,
        y = ~MATHS,
        color = ~RACE)
```

### The Code

```{r}
#| eval: false
#| fig-widith: 12
#| fig-height: 8
plot_ly(data = exam_data,
        x = ~ENGLISH,
        y = ~MATHS,
        color = ~RACE)
```
:::

### Coordinated Multiple views with plotly

The creation of a coordinated linked plot by using plotly involves three steps:

- highlight_key() of plotly package is used as shared data. two scatterplots will be created by using ggplot2 functions. lastly, subplot() of plotly package is used to place them next to each other side-by-side.

- Lesson learned from the code chunk: highlight_key() simply creates an object of class [crosstalk::SharedData.](https://rdrr.io/cran/crosstalk/man/SharedData.html)

::: panel-tabset
### The Plot

```{r}
#| echo: false
#| fig-widith: 12
#| fig-height: 8
d <- highlight_key(exam_data)
p1 <- ggplot(data=d,
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
p2 <- ggplot(data=d,
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```

### The Code

```{r}
#| eval: false
#| fig-widith: 12
#| fig-height: 8
d <- highlight_key(exam_data)
p1 <- ggplot(data=d,
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
p2 <- ggplot(data=d,
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```
:::

### Interactive data visualisation - crosstalk methods

[Crosstalk](https://rstudio.github.io/crosstalk/index.html) is an add-on to the htmlwidgets package. It extends htmlwidgets with a set of classes, functions, and conventions for implementing cross-widget interactions (currently, linked brushing and filtering).

### Interactive data table: DT package

- A wrapper of the JavaScript Library [DataTables](https://datatables.net/)

- Data objects in R can be rendered as HTML tables using the JavaScript library ‘DataTables’ (typically via R Markdown or Shiny).

```{r}
#| fig-widith: 12
#| fig-height: 8
DT::datatable(exam_data, class= "compact")
```

### Linked brushing: crosstalk method

```{r}
#| fig-widith: 12
#| fig-height: 8
d <- highlight_key(exam_data)
p <- ggplot(d,
            aes(ENGLISH,
                MATHS)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
gg <- highlight(ggplotly(p),
                "plotly_selected")
crosstalk::bscols(gg,
                  DT::datatable(d),
                  widths = 5)
```

## Chapter 4: Programming Animated Statistical Graphics with R

### Install package

- [**plotly**](https://plotly.com/r/), R library for plotting interactive statistical graphs.

- [**gganimate**](https://gganimate.com/), an ggplot extension for creating animated statistical graphs.

- [**gifski**](https://cran.r-project.org/web/packages/gifski/index.html) converts video frames to GIF animations using pngquant’s fancy features for efficient cross-frame palettes and temporal dithering. It produces animated GIFs that use thousands of colors per frame.

- [**gapminder**](https://cran.r-project.org/web/packages/gapminder/index.html): An excerpt of the data available at Gapminder.org. We just want to use its *country_colors* scheme.

- [**tidyverse**](https://www.tidyverse.org/), a family of modern R packages specially designed to support data science, analysis and communication task including creating static statistical graphs.

```{r}
pacman::p_load(readxl, gifski, gapminder,
               plotly, gganimate, tidyverse)
```

### Import Data

```{r}
col <- c("Country", "Continent")
globalPop <- read_xls("data/GlobalPopulation.xls",
                      sheet="Data") %>%
  mutate_at(col, as.factor) %>%
  mutate(Year = as.integer(Year))
glimpse(globalPop)
```

### Animated Data Visualisation: gganimate methods

gganimate extends the grammar of graphics as implemented by ggplot2 to include the description of animation. It does this by providing a range of new grammar classes that can be added to the plot object in order to customise how it should change with time.

- `transition_*()` defines how the data should be spread out and how it relates to itself across time.\*

- `view_*()` *defines how the positional scales should change along the animation.*

- `shadow_*()` *defines how data from other points in time should be presented in the given point in time.*

- `enter_*()/exit_*()` *defines how new data should appear and how old data should disappear during the course of the animation.*

- `ease_aes()` *defines how different aesthetics should be eased during transitions.*

### Building a static population bubble plot

::: panel-tabset
### The Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
ggplot(globalPop, aes(x = Old, y = Young,
                      size = Population,
                      colour = Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}',
       x = '% Aged',
       y = '% Young')
```

### The Code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
ggplot(globalPop, aes(x = Old, y = Young,
                      size = Population,
                      colour = Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}',
       x = '% Aged',
       y = '% Young')
```
:::

### Building a animated bubble plot

- [`transition_time()`](https://gganimate.com/reference/transition_time.html) of **gganimate** is used to create transition through distinct states in time (i.e. Year).

- `ease_aes()` is used to control easing of aesthetics. The default is `linear`. Other methods are: quadratic, cubic, quartic, quintic, sine, circular, exponential, elastic, back, and bounce.

::: panel-tabset
### The Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
ggplot(globalPop, aes(x = Old, y = Young,
                      size = Population,
                      colour = Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}',
       x = '% Aged',
       y = '% Young') +
  transition_time(Year) +
  ease_aes('linear')
```

### The Code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
ggplot(globalPop, aes(x = Old, y = Young,
                      size = Population,
                      colour = Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}',
       x = '% Aged',
       y = '% Young') +
  transition_time(Year) +
  ease_aes('linear')
```
:::

### Animated data visualisation: plotly()

In **Plotly R** package, both `ggplotly()` and `plot_ly()` support key frame animations through the `frame` argument/aesthetic. They also support an `ids` argument/aesthetic to ensure smooth transitions between objects with the same id (which helps facilitate object constancy).

### Building an animated bubble plot: ggplotly() method

::: panel-tabset
### The Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
gg <- ggplot(globalPop,
       aes(x = Old,
           y = Young,
           size = Population,
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged',
       y = '% Young')
ggplotly(gg)
```

### The Code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
gg <- ggplot(globalPop,
       aes(x = Old,
           y = Young,
           size = Population,
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged',
       y = '% Young')
ggplotly(gg)
```
:::

To remove the legend and provide better visualisation, `theme(legend.position='none')` should be used as shown in the plot and code chunk below.

::: panel-tabset
### The Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
gg <- ggplot(globalPop,
       aes(x = Old,
           y = Young,
           size = Population,
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged',
       y = '% Young') +
  theme(legend.position='none')
ggplotly(gg)
```

### The Code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
gg <- ggplot(globalPop,
       aes(x = Old,
           y = Young,
           size = Population,
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged',
       y = '% Young') +
  theme(legend.position='none')
ggplotly(gg)
```
:::

### Building an animated bubble plot: plot_ly() method

::: panel-tabset
### The Plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
bp <- globalPop %>%
  plot_ly(x = ~Old,
          y = ~Young,
          size = ~Population,
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year,
          text = ~Country,
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers'
          ) %>%
  layout(showlegend = FALSE)
bp
```

### The Code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
bp <- globalPop %>%
  plot_ly(x = ~Old,
          y = ~Young,
          size = ~Population,
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year,
          text = ~Country,
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers'
          ) %>%
  layout(showlegend = FALSE)
bp
```
:::
