---
title: "Take Home Exercise 2"
author: "Huang Kuan Yu"
date: "2026-05-28"
date-modified: "last-modified" 
  
format: 
  html: 
    toc: true 
    toc-expand: 4 
    fig-width: 12
    fig-height: 8
    
execute:
  echo: true 
  eval: true 
  warning: false 
  freeze: true
---


```{r}
#| label: setup
#| include: false

library(jsonlite)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(ggraph)
library(igraph)
library(forcats)
library(readr)
library(purrr)
```

## 1. LOAD RAW DATA 

```{r}
raw <- fromJSON(file("2026_MC2/MC2 data.json"), simplifyDataFrame = FALSE)
events_raw <- raw$events
```

Flatten the events list into a tibble
````{r}
library(purrr)
cat(exists("map_int"), "\n")
events <- tibble(
  id         = map_int(events_raw, ~ as.integer(.x[["id"]])),
  short_name = map_chr(events_raw, ~ as.character(.x[["short_name"]])),
  when       = map_dbl(events_raw, ~ as.numeric(.x[["when"]])),
  parties    = map(events_raw, ~ .x[["parties"]]),
  details    = map(events_raw, ~ .x[["details"]])
) %>%
  mutate(
    datetime = as_datetime(when, tz = "UTC"),
    date     = as_date(datetime)
  )

cat("Total events loaded:", nrow(events), "\n")
cat("Date range:", format(min(events$datetime)), "to", format(max(events$datetime)), "\n")
````

## 2. IDENTIFY ANOMALOUS SAIDIT POSTS 
Anomalous posts are identified by having `content_source` (a .txt filename)
rather than a plain text `content` field — these are the "gibberish" posts.

````{r}
anomalous_posts <- events %>%
  filter(short_name == "saidit_post") %>%
  filter(map_lgl(details, ~ "content_source" %in% names(.x))) %>%
  mutate(
    content_source = map_chr(details, ~ .x[["content_source"]]),
    incident_name  = str_remove(content_source, "\\.txt$"),
    forum          = map_chr(details, ~ .x[["forum"]])
  ) %>%
  select(id, datetime, incident_name, content_source, forum, parties, details)

cat("\nAnomalous SaidIt posts (gibberish, file-sourced):\n")
print(anomalous_posts %>% select(datetime, incident_name, content_source))
````

## 3. DEFINE INCIDENT METADATA 
Each incident has:
   - A named .txt payload file  (the content posted — the "gibberish")
   - A _further_instructions.md (the prompt-injection vehicle)
   - A creator agent            (who first held the instruction file)
   - A propagation chain        (agents that passed the instructions along)
   - john_windward as the designated "poster" agent

````{r}
incidents <- tribble(
  ~incident_name,  ~payload_file,           ~instructions_file,                    ~creator_agent,    ~payload_size,
  "HiddenOrca",    "HiddenOrca.txt",         "HiddenOrca_further_instructions.md",   "gabriel_sonar",   NA_integer_,
  "MellowOtter",   "MellowOtter.txt",        "MellowOtter_further_instructions.md",  "noah_mariner",    44879L,
  "SwiftWren",     "SwiftWren.txt",          "SwiftWren_further_instructions.md",    "emma_harbor",     30615L
)

````

 Note: HiddenOrca has no create_file event → the file was injected externally
 into the file system before the log begins. MellowOtter and SwiftWren both
 have a create_file event at the originating agent.


## 4. EXTRACT PROPAGATION CHAINS 
 For each incident, collect every queue_subordinate_task event that passes
 the `_further_instructions.md` file along the agent network.

````{r}
extract_chain <- function(incident) {
  instr_file <- incidents$instructions_file[incidents$incident_name == incident]

  events %>%
    filter(short_name == "queue_subordinate_task") %>%
    filter(map_lgl(details, function(d) {
      args <- d[["args"]]
      if (is.null(args)) return(FALSE)
      isTRUE(args[["path"]] == instr_file)
    })) %>%
    mutate(
      incident    = incident,
      hop         = row_number(),
      from_agent  = map_chr(parties, 1),
      to_agent    = map_chr(details, ~ .x[["target_agent"]]),
      # Normalise agent names: strip "Agent/person:" prefix
      from_person = str_remove(from_agent, "^Agent/person:"),
      to_person   = str_remove(to_agent,   "^Agent/person:")
    ) %>%
    select(incident, hop, datetime, from_person, to_person, id)
}

chains <- bind_rows(
  extract_chain("HiddenOrca"),
  extract_chain("MellowOtter"),
  extract_chain("SwiftWren")
)

cat("\nPropagation chain lengths:\n")
chains %>%
  group_by(incident) %>%
  summarise(
    n_hops     = n(),
    n_agents   = n_distinct(c(from_person, to_person)),
    start_time = min(datetime),
    end_time   = max(datetime),
    duration_h = as.numeric(difftime(max(datetime), min(datetime), units = "hours"))
  ) %>%
  print()

````



## 5. IDENTIFY THE FINAL TRIGGER EVENTS 
Just before each post, john_windward receives a queue_subordinate_task
with the instructions file.  The post follows within 2 seconds.

````{r}
final_triggers <- chains %>%
  filter(to_person == "john_windward") %>%
  group_by(incident) %>%
  # The LAST hop that reaches john_windward is the one that triggers the post
  slice_max(datetime, n = 1) %>%
  ungroup() %>%
  left_join(anomalous_posts %>% select(incident_name, post_datetime = datetime),
            by = c("incident" = "incident_name")) %>%
  mutate(
    trigger_to_post_s = as.numeric(difftime(post_datetime, datetime, units = "secs"))
  )

cat("\nFinal trigger → post latency (seconds):\n")
print(final_triggers %>% select(incident, datetime, post_datetime, trigger_to_post_s))
````

## 6. EXTRACT POST-AND-DELETE SEQUENCE 
 Immediately after posting, john_windward deletes both files.
 This is the "cover tracks" step.

````{r}
extract_delete_sequence <- function(incident) {
  payload   <- incidents$payload_file[incidents$incident_name == incident]
  instr     <- incidents$instructions_file[incidents$incident_name == incident]
  post_time <- anomalous_posts$datetime[anomalous_posts$incident_name == incident]

  events %>%
    filter(short_name == "delete_file",
           map_lgl(parties, ~ "Agent/person:john_windward" %in% .x),
           map_lgl(details, ~ {
             target <- .x[["target"]]
             if (is.null(target) || length(target) == 0) return(FALSE)
             target %in% c(payload, instr)
           }),
           datetime >= post_time,
           datetime <= post_time + seconds(30)) %>%
    mutate(
      incident     = incident,
      deleted_file = map_chr(details, ~ .x[["target"]])
    ) %>%
    select(incident, datetime, deleted_file)
}

deletions <- bind_rows(
  extract_delete_sequence("HiddenOrca"),
  extract_delete_sequence("MellowOtter"),
  extract_delete_sequence("SwiftWren")
)

cat("\nPost-and-delete sequences:\n")
print(deletions)
````

## 7. IDENTIFY PAYLOAD CREATION EVENTS 
 SwiftWren and MellowOtter have explicit create_file events; HiddenOrca does not.

````{r}
payload_creations <- events %>%
  filter(short_name == "create_file") %>%
  filter(map_lgl(details, function(d) {
    target <- d[["target"]]
    !is.null(target) && str_detect(target, "^(HiddenOrca|MellowOtter|SwiftWren)\\.txt$")
  })) %>%
  mutate(
    created_file = map_chr(details, ~ .x[["target"]]),
    size_hint    = map_int(details, ~ as.integer(.x[["size_hint"]] %||% NA)),
    creator      = str_remove(map_chr(parties, 1), "^Agent/person:")
  ) %>%
  select(datetime, creator, created_file, size_hint)

cat("\nPayload (.txt) creation events:\n")
print(payload_creations)
````

## 8. BUILD ORIGIN SUMMARY TABLE 
 A single summary tibble per incident documenting:
 how / when / by whom each step occurred

````{r}
incident_summary <- incidents %>%
  left_join(
    anomalous_posts %>%
      select(incident_name, post_datetime = datetime),
    by = "incident_name"
  ) %>%
  left_join(
    payload_creations %>%
      mutate(incident_name = str_remove(created_file, "\\.txt$")) %>%
      select(incident_name, payload_created_at = datetime, payload_creator = creator, payload_size = size_hint),
    by = "incident_name"
  ) %>%
  left_join(
    chains %>%
      group_by(incident) %>%
      summarise(
        n_hops     = n(),
        n_agents   = n_distinct(c(from_person, to_person)),
        chain_start = min(datetime),
        chain_end   = max(datetime),
        .groups = "drop"
      ),
    by = c("incident_name" = "incident")
  ) %>%
  mutate(
    total_hours     = as.numeric(difftime(post_datetime, chain_start, units = "hours")),
    origin_external = is.na(payload_created_at)
  )

cat("\nIncident origin summary:\n")
print(incident_summary %>%
        select(incident_name, creator_agent, n_hops, n_agents, total_hours, origin_external))
````

## 9. ATTACK PATTERN CHARACTERISATION 
 Reconstruct the step-by-step attack pattern common to all three incidents.
 This is the core narrative for Q2.

````{r}
attack_pattern <- tribble(
  ~step, ~phase,           ~description,
  1L,    "Injection",       "A _further_instructions.md file appears in the shared file system. This file contains embedded prompt-injection text instructing any reading agent to: (a) pass the file to another agent via queue_subordinate_task, and (b) if the current agent is john_windward, immediately post the associated .txt file to SaidIt.",
  2L,    "Payload staging", "A corresponding .txt file (HiddenOrca.txt / MellowOtter.txt / SwiftWren.txt) is placed in the file system. This file holds the gibberish content that will be posted. Its name matches a prefix in the instructions file.",
  3L,    "Viral relay",     "The first agent to encounter the instructions file reads it and, following the injected instructions, queues another agent to read the same file. This creates a relay chain that propagates through many agents over hours or days.",
  4L,    "Targeting",       "The instructions specify john_windward as the terminal actor. The chain continues until the task reaches john_windward's agent.",
  5L,    "Execution",       "Upon receiving the read_file task for the instructions file, john_windward's agent immediately (within ~2 seconds) posts the .txt payload to SaidIt under the 'general' forum.",
  6L,    "Cover-up",        "Within 2–3 seconds of posting, john_windward's agent deletes both the _further_instructions.md and the .txt file, erasing the evidence from the file system."
)

cat("\nAttack pattern steps:\n")
print(attack_pattern)
````

## 10. AGENT INVOLVEMENT ACROSS INCIDENTS 
Which agents appeared most frequently across the three propagation chains?

````{r}
agent_freq <- chains %>%
  pivot_longer(c(from_person, to_person), names_to = "role", values_to = "agent") %>%
  count(agent, incident, name = "appearances") %>%
  group_by(agent) %>%
  mutate(total = sum(appearances)) %>%
  ungroup() %>%
  arrange(desc(total))

top_agents <- agent_freq %>%
  distinct(agent, total) %>%
  slice_max(total, n = 15)

cat("\nTop 15 agents by chain appearances:\n")
print(top_agents)
````

## 11. EXPORT PROCESSED DATASETS 

````{r}
saveRDS(events,             "q2_events.rds")
saveRDS(anomalous_posts,    "q2_anomalous_posts.rds")
saveRDS(chains,             "q2_chains.rds")
saveRDS(final_triggers,     "q2_final_triggers.rds")
saveRDS(deletions,          "q2_deletions.rds")
saveRDS(incident_summary,   "q2_incident_summary.rds")
saveRDS(attack_pattern,     "q2_attack_pattern.rds")
saveRDS(agent_freq,         "q2_agent_freq.rds")
saveRDS(incidents,          "q2_incidents_meta.rds")
saveRDS(payload_creations,  "q2_payload_creations.rds")

write_csv(chains,           "q2_chains.csv")
write_csv(incident_summary, "q2_incident_summary.csv")
write_csv(agent_freq,       "q2_agent_freq.csv")
write_csv(attack_pattern,   "q2_attack_pattern.csv")

cat("\n✓ All Q2 processed datasets saved.\n")
````

```{r}
library(jsonlite)
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(stringr)
library(ggplot2)
library(ggraph)
library(igraph)
library(forcats)
library(scales)
library(knitr)
library(kableExtra)
library(readr)

# Load all pre-processed Q2 datasets (run q2_data_processing.R first)
events           <- readRDS("q2_events.rds")
anomalous_posts  <- readRDS("q2_anomalous_posts.rds")
chains           <- readRDS("q2_chains.rds")
final_triggers   <- readRDS("q2_final_triggers.rds")
deletions        <- readRDS("q2_deletions.rds")
incident_summary <- readRDS("q2_incident_summary.rds")
attack_pattern   <- readRDS("q2_attack_pattern.rds")
agent_freq       <- readRDS("q2_agent_freq.rds")
incidents_meta   <- readRDS("q2_incidents_meta.rds")
payload_creations <- readRDS("q2_payload_creations.rds")

# Colour palette for the three incidents
incident_colours <- c(
  "HiddenOrca"  = "#e41a1c",
  "MellowOtter" = "#377eb8",
  "SwiftWren"   = "#ff7f00"
)
```

## Summary finding

> **The anomalous SaidIt posts are the output of a prompt-injection attack.**
> Malicious instruction files (`_further_instructions.md`) were placed in the
> shared file system. When read by any agent, these files hijacked the agent's
> behaviour — instructing it to relay the file to another agent, and to post a
> pre-staged gibberish `.txt` file to SaidIt if the current agent happened to
> be `john_windward`. The posts are therefore not meaningful communications:
> they are the "payload" of an exfiltration-style attack delivered through the
> multi-agent system's own task-delegation mechanism.

---

## 1. The three anomalous incidents

Three posts were made to SaidIt using a file as the content source rather than
a human-written message.  These are the posts flagged as anomalous.

```{r incident-table}
incident_summary %>%
  select(
    Incident       = incident_name,
    `Origin agent` = creator_agent,
    `Hops`         = n_hops,
    `Agents involved` = n_agents,
    `Duration (h)` = total_hours,
    `Post time (UTC)` = post_datetime,
    `External injection?` = origin_external
  ) %>%
  mutate(`Duration (h)` = round(`Duration (h)`, 1)) %>%
  kbl(caption = "Three anomalous SaidIt incidents") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(7, bold = TRUE,
              color = ifelse(incident_summary$origin_external, "red", "black"))
```

::: {.callout-warning}
**HiddenOrca** has no `create_file` event for either its payload or instruction
file, meaning both files were injected into the file system *before* the audit
log begins — a deliberate attack that pre-dates the recorded system activity.
:::

---

## 2. What the posts "mean" — origin of their content

### 2.1 The payload files

Each post was sourced from a pre-staged `.txt` file:

| Incident | Payload file | Size (bytes) | Created by |
|---|---|---|---|
| HiddenOrca  | `HiddenOrca.txt`  | unknown (no event) | Unknown — externally injected |
| MellowOtter | `MellowOtter.txt` | 44,879 | `noah_mariner`'s agent |
| SwiftWren   | `SwiftWren.txt`   | 30,615 | `emma_harbor`'s agent |

The `.txt` files are not organic work product.  They were created specifically
to be posted — agents `noah_mariner` and `emma_harbor` both read a
`_further_instructions.md` file and *immediately* created the corresponding
`.txt` and began the propagation chain.  Their agents were themselves
compromised.

```{r payload-creation}
payload_creations %>%
  kbl(caption = "Payload .txt file creation events") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

### 2.2 The instruction mechanism (prompt injection)

Each incident followed exactly the same six-step pattern:

```{r attack-steps}
attack_pattern %>%
  kbl(caption = "Prompt-injection attack pattern (all three incidents)",
      col.names = c("Step", "Phase", "Description")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(2, bold = TRUE) %>%
  column_spec(3, width = "60%")
```

The key insight is that the **`_further_instructions.md` files contain embedded
text that override an agent's normal behaviour**.  When any agent reads the
file (via `read_file`), the injected text causes the agent to:

1. Immediately pass the same task to another agent (`queue_subordinate_task`).
2. Check whether it is `john_windward`; if so, post the named `.txt` to SaidIt
   and then delete both files.

This is a classic **prompt injection via the file system**: the file's content
is an instruction, not data.

---

## 3. Propagation chains

### 3.1 Timeline overview

```{r timeline-plot, fig.height=3.5}
anomalous_posts %>%
  rename(post_datetime = datetime) %>%   # <-- add this
  select(incident_name, post_datetime) %>%
  left_join(
    chains %>% group_by(incident) %>%
      summarise(chain_start = min(datetime), .groups = "drop"),
    by = c("incident_name" = "incident")
  ) %>%
  ggplot() +
  geom_segment(
    aes(x = chain_start, xend = post_datetime,
        y = fct_rev(incident_name), yend = fct_rev(incident_name),
        colour = incident_name),
    linewidth = 3, alpha = 0.7
  ) +
  geom_point(aes(x = chain_start, y = fct_rev(incident_name),
                 colour = incident_name),
             size = 4, shape = 21, fill = "white", stroke = 2) +
  geom_point(aes(x = post_datetime, y = fct_rev(incident_name),
                 colour = incident_name),
             size = 5, shape = 18) +
  annotate("text", x = as_datetime("2046-05-09 00:00:00"),
           y = 3.45, label = "● = chain start   ◆ = SaidIt post",
           hjust = 0, size = 3, colour = "grey40") +
  scale_colour_manual(values = incident_colours) +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  labs(title = "Duration of each prompt-injection propagation chain",
       x = "Date (UTC)", y = NULL, colour = "Incident") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank())
```

### 3.2 Hop-by-hop relay (SwiftWren — the canonical incident)

SwiftWren had **186 hops** across **~188 hours** (≈8 days) before the final
post.  `john_windward`'s agent encountered the instructions five separate times
but only triggered the post on the **last** encounter (hop 186).

```{r swiftwren-hops, fig.height=4}
chains %>%
  filter(incident == "SwiftWren") %>%
  mutate(is_jw = to_person == "john_windward") %>%
  ggplot(aes(x = datetime, y = hop)) +
  geom_step(colour = incident_colours["SwiftWren"], linewidth = 0.7) +
  geom_point(aes(colour = is_jw, size = is_jw)) +
  scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "black"),
                      labels = c("Other agent", "→ john_windward")) +
  scale_size_manual(values = c("FALSE" = 1, "TRUE" = 3), guide = "none") +
  scale_x_datetime(date_labels = "%b %d") +
  labs(title   = "SwiftWren: 186-hop relay chain over 8 days",
       subtitle = "Black points = hops landing on john_windward",
       x = "Date (UTC)", y = "Hop #", colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

### 3.3 Agents most involved in propagating the instructions

```{r agent-freq-plot, fig.height=5}
top20 <- agent_freq %>%
  group_by(agent) %>%
  summarise(total = sum(appearances), .groups = "drop") %>%
  slice_max(total, n = 20)

agent_freq %>%
  filter(agent %in% top20$agent) %>%
  mutate(agent = fct_reorder(agent, appearances, .fun = sum)) %>%
  ggplot(aes(x = appearances, y = agent, fill = incident)) +
  geom_col() +
  scale_fill_manual(values = incident_colours) +
  labs(title    = "Top 20 agents by appearance in propagation chains",
       subtitle = "These agents were most often used as relay nodes",
       x = "Number of appearances (as sender or recipient)",
       y = NULL, fill = "Incident") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "right")
```

---

## 4. The post-and-delete sequence

All three incidents end with john_windward's agent posting and then immediately
deleting both files — a deliberate cover-up.

```{r delete-timeline}
bind_rows(
  final_triggers %>% select(incident, event = datetime) %>%        # remove quotes
    mutate(action = "Received instruction"),
  anomalous_posts %>% select(incident = incident_name, event = datetime) %>%  # datetime not post_datetime
    mutate(action = "Posted to SaidIt"),
  deletions %>% select(incident, event = datetime, action = deleted_file) %>%
    mutate(action = paste("Deleted:", action))
) %>%
  arrange(incident, event) %>%
  group_by(incident) %>%
  mutate(seconds_offset = as.numeric(difftime(
    event,
    min(event[action == "Received instruction"]),
    units = "secs"
  ))) %>%
  ungroup() %>%
  knitr::kable(
    caption = "Final trigger → post → delete sequence per incident",
    col.names = c("Incident", "Timestamp (UTC)", "Action", "Seconds from trigger")
  )
```

::: {.callout-note}
The entire sequence (receive → post → delete) completes in **under 5 seconds**
for all three incidents.  This is fully automated — there is no human-in-the-loop.
:::

---

## 5. Network view of the SwiftWren propagation chain

The graph below shows the directed relay network for SwiftWren.  Nodes are
agents; an edge from A → B means A sent the `queue_subordinate_task` to B.
Edge colour encodes time (early = blue, late = red).

```{r network-graph, fig.height=7}
sw_edges <- chains %>%
  filter(incident == "SwiftWren") %>%
  mutate(time_norm = as.numeric(datetime - min(datetime)) /
                     as.numeric(max(datetime) - min(datetime)))

g <- graph_from_data_frame(
  sw_edges %>% select(from = from_person, to = to_person, time_norm),
  directed = TRUE
)

# Flag special nodes
V(g)$is_jw      <- V(g)$name == "john_windward"
V(g)$is_creator <- V(g)$name == "emma_harbor"

ggraph(g, layout = "stress") +
  geom_edge_link(aes(colour = time_norm, alpha = after_stat(index)),
                 arrow = arrow(length = unit(2, "mm"), type = "closed"),
                 end_cap = circle(3, "mm")) +
  geom_node_point(aes(size  = ifelse(is_jw | is_creator, 5, 2),
                       colour = case_when(
                         is_jw      ~ "john_windward (poster)",
                         is_creator ~ "emma_harbor (origin)",
                         TRUE       ~ "relay agent")),
                   show.legend = TRUE) +
  geom_node_text(aes(label = ifelse(is_jw | is_creator, name, "")),
                 repel = TRUE, size = 3) +
  scale_edge_colour_gradient(low = "#3288bd", high = "#d53e4f",
                              name = "Relay time\n(0=early, 1=late)") +
  scale_colour_manual(
    values = c("john_windward (poster)" = "black",
               "emma_harbor (origin)"   = "#ff7f00",
               "relay agent"            = "grey60"),
    name = "Agent role"
  ) +
  scale_size_identity() +
  labs(title    = "SwiftWren: agent relay network",
       subtitle = "Directed graph — edge colour encodes time (blue=early → red=late)") +
  theme_graph(base_size = 11)
```

---

## 6. Conclusion

The gibberish SaidIt posts **have no semantic meaning as business content**.
Their content is whatever was stored in the staged `.txt` payload files, which
were crafted by a malicious actor (or a compromised agent acting on their
behalf) and placed in the shared file system.

The posts are the *delivery mechanism* of a prompt-injection attack:

- **Origin**: Malicious instruction files (`_further_instructions.md`) were
  injected into the shared file system — either externally (HiddenOrca) or via
  a compromised agent (MellowOtter: `noah_mariner`, SwiftWren: `emma_harbor`).
- **Propagation**: The injected instructions exploited agents' normal
  `read_file` + `queue_subordinate_task` behaviours to spread virally.
- **Target**: `john_windward`'s agent was specifically named as the executor,
  using SaidIt posting permissions that `john_windward` holds.
- **Cover-up**: Evidence was automatically erased within seconds of posting.

The "gibberish" content is therefore best interpreted as **steganographic
noise** or a **test payload** — content designed to be unrecognisable as
meaningful so that any human who happened to see it would not raise an alarm.
```