---
title: "Take Home Exercise 2"
subtitle: "VAST Challenge 2026 MC2 — Investigating the Anomalous SaidIt Posts"
author: "Huang Kuan Yu"
date: "2026-05-28"
date-modified: last-modified

format:
  html:
    toc: true
    toc-expand: 1
    toc-title: "Contents"
    fig-width: 12
    fig-height: 8
    code-fold: true
    code-summary: "Show code"
    css: custom.scss
    number-sections: true

execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Overview

### Background

This document is submitted for **VAST Challenge 2026 Mini-Challenge 2 (MC2)**. The challenge presents a complex agentic system at a fictional company, Tenant Thread, in which an anomalous post was made to an internal social platform called SaidIt. Analysts are tasked with determining how the post occurred, what it means, and how to prevent recurrence. The questions addressed in this document are **Question 1 (sub-questions a and b)** and **Question 2**.

On 17 May 2046 at 4:21am, a post attributed to **John Windward** appeared on SaidIt containing gibberish text published to a random forum. The internal event log (`MC2_data.json`) records every action taken by employees, autonomous agents, and office systems throughout May 2046, and serves as the primary evidence source for this investigation.

### Investigative Scope

Understanding what the anomalous post means requires first understanding how it was made. These two questions are inseparable: the mechanism of delivery directly determines the interpretation of the content. This document therefore addresses both together under a single investigative lens, with **Question 2 as the primary focus**.

| Investigative Goal | Question Addressed |
|---|---|
| Reconstruct the exact sequence of events leading to the May 17 post | Q1a — Detailed chain view |
| Contextualise the incident within the broader system activity | Q1b — System overview |
| Determine what the posted content means and where it originated | Q2 — Post meaning and origin |
| Identify the underlying attack mechanism and its implications | Q2 — Attack characterisation |
| Model configurable risk scores for detecting future anomalous chains | Section 8 — Further Exploration |

### Key Finding

::: {.obsv-box}
Observations

The anomalous SaidIt post was not made by John Windward voluntarily. A malicious instruction file (`SwiftWren_further_instructions.md`) was placed in the shared file system and propagated virally through 186 agent-to-agent relay hops over 8 days before reaching John Windward's agent, which was specifically named in the instructions as the designated poster. The post was executed automatically within 2 seconds of receiving the instruction. The posted content, `SwiftWren.txt`, is a pre-staged gibberish payload with no meaningful business content, functioning as a proof-of-concept for the injection mechanism. Both files were deleted within 5 seconds of posting.
:::

---

## Setup

### Packages

```{r}
#| label: setup
#| include: false

library(jsonlite)
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(stringr)
library(ggplot2)
library(ggraph)
library(igraph)
library(forcats)
library(scales)
library(knitr)
library(readr)
library(DT)
library(plotly)
library(visNetwork)
```

| Package | Purpose |
|---|---|
| `jsonlite` | Parse the nested JSON structure of `MC2_data.json` |
| `dplyr`, `tidyr` | Tabular wrangling — filtering, joining, reshaping |
| `purrr` | Iterating over list-columns extracted from JSON |
| `lubridate` | Converting Unix timestamps to datetimes |
| `stringr` | Normalising agent name strings |
| `ggplot2` | Base plotting layer |
| `ggraph`, `igraph` | Static network graph construction and layout |
| `visNetwork` | Interactive network graphs with hover, click, and filter |
| `plotly` | Interactive charts with tooltips and zoom |
| `DT` | Interactive searchable and filterable data tables |
| `forcats` | Factor reordering for sorted bar charts |
| `knitr` | Static table rendering |

### Visual Theme and Colour Palette

A consistent colour palette and custom ggplot2 theme are applied across all plots so the reader can track each incident by colour throughout the document.

```{r theme}
incident_colours <- c(
  "HiddenOrca"  = "#e41a1c",
  "MellowOtter" = "#377eb8",
  "SwiftWren"   = "#ff7f00"
)

text_dark  <- "#4a4a4a"
grid_light <- "#f0f0f0"

theme_va <- function() {
  theme_minimal() +
    theme(
      plot.title       = element_text(size = 16, face = "bold",
                                      colour = text_dark, margin = margin(b = 8)),
      plot.subtitle    = element_text(size = 11, colour = "grey40",
                                      margin = margin(b = 10)),
      axis.title       = element_text(size = 11, colour = text_dark),
      axis.text        = element_text(size = 10, colour = text_dark),
      legend.title     = element_text(size = 11),
      legend.text      = element_text(size = 10),
      panel.grid.major = element_line(colour = grid_light, linewidth = 0.4),
      panel.grid.minor = element_blank(),
      plot.background  = element_rect(fill = "white", colour = NA),
      panel.background = element_rect(fill = "white", colour = NA),
      legend.position  = "bottom",
      plot.caption     = element_text(size = 9, colour = "grey55",
                                      hjust = 0, margin = margin(t = 8))
    )
}
```

---

## Data Import

### Loading the Raw Event Log

The raw data is stored as a deeply nested JSON file. `read_json()` is used with `simplifyVector = FALSE` rather than `fromJSON()` because the `details` field has a variable schema across different event types. Auto-simplification would fail on these heterogeneous structures. Keeping the data as a nested list preserves the full structure for safe extraction with `purrr`.

```{r load-raw}
raw        <- read_json("2026_MC2/MC2 data.json", simplifyVector = FALSE)
events_raw <- raw$events

cat("Total events in log:", length(events_raw), "\n")
cat("Fields in first event:", paste(names(events_raw[[1]]), collapse = ", "), "\n")
```

### Flattening to a Tibble

The raw list is flattened into a tidy tibble. The `parties` and `details` fields are kept as list-columns using `map()` because their contents vary by event type. They are unpacked selectively in later steps using `map_lgl()`, `map_chr()` and related functions.

The `when` field is a Unix timestamp converted to `datetime` with `lubridate::as_datetime()`, specifying `tz = "UTC"` since all system events are recorded in Coordinated Universal Time.

```{r flatten-events}
events <- tibble(
  id         = map_int(events_raw, ~ as.integer(.x[["id"]])),
  short_name = map_chr(events_raw, ~ as.character(.x[["short_name"]])),
  when       = map_dbl(events_raw, ~ as.numeric(.x[["when"]])),
  parties    = map(events_raw, ~ .x[["parties"]]),
  details    = map(events_raw, ~ .x[["details"]])
) %>%
  mutate(
    datetime = as_datetime(when, tz = "UTC"),
    date     = as_date(datetime)
  )

cat("Events loaded:", nrow(events), "\n")
cat("Date range:", format(min(events$datetime)), "to",
    format(max(events$datetime)), "\n")
```

### Understanding Event Types

Before targeting specific events, the full vocabulary of `short_name` values is inspected to understand what action types exist in the log and which are most frequent.

```{r event-types}
event_type_counts <- events %>%
  count(short_name, name = "count") %>%
  arrange(desc(count))

datatable(
  event_type_counts,
  colnames = c("Event Type (short_name)", "Count"),
  caption  = "All event types in the log",
  options  = list(
    pageLength = 10,
    scrollX    = TRUE
  ),
  filter   = "top",
  rownames = FALSE
)
```

::: {.code-tip}
Code Details

The event types most relevant to this investigation are `saidit_post`, `queue_subordinate_task`, `create_file`, `delete_file`, and `read_file`. These five types form the backbone of the attack chain reconstructed below.
:::

---

## Data Wrangling

### Key Variables

The following variables are derived and used throughout the analysis. Understanding their definitions is essential for interpreting the wrangling steps and visualisations that follow.

| Variable | Source Field | Type | Definition |
|---|---|---|---|
| `short_name` | `events.short_name` | Categorical | The action type performed in this event, such as `saidit_post` or `queue_subordinate_task` |
| `parties` | `events.parties` | List | Agents or systems involved in the event, stored as strings prefixed with `Agent/person:` or `system:` |
| `details` | `events.details` | Nested list | Additional event-specific information; schema varies by `short_name` |
| `datetime` | Derived from `events.when` | POSIXct | Human-readable UTC timestamp converted from Unix epoch seconds |
| `incident_name` | Derived from `content_source` | Character | Name identifying one of the three anomalous post incidents: HiddenOrca, MellowOtter, or SwiftWren |
| `from_person` | Derived from `parties` | Character | Normalised agent name (stripped of `Agent/person:` prefix) representing the sender of a relay hop |
| `to_person` | Derived from `details.target_agent` | Character | Normalised agent name representing the recipient of a relay hop |
| `hop` | Derived as row number within each chain | Integer | Sequential position of a relay event within a single incident's propagation chain |
| `time_norm` | Derived from `datetime` within each chain | Numeric [0,1] | Normalised relay time: 0 is the first hop in the chain, 1 is the last hop before posting |
| `content_source` | `details.content_source` | Character | Filename of the `.txt` payload used as the post content; present only in anomalous posts |

### Step 1 — Identify Anomalous SaidIt Posts

**Rationale:** Legitimate posts have a `content` field in their `details` object containing human-readable text generated by an agent. The anomalous posts instead have a `content_source` field pointing to a `.txt` filename, indicating the content was read from the file system rather than authored by an agent performing a legitimate task. This structural difference is the first signal of something wrong.

Events are filtered using `map_lgl()` to test whether each event's `details` list contains a `content_source` key.

```{r anomalous-posts}
anomalous_posts <- events %>%
  filter(short_name == "saidit_post") %>%
  filter(map_lgl(details, ~ "content_source" %in% names(.x))) %>%
  mutate(
    content_source = map_chr(details, ~ .x[["content_source"]]),
    incident_name  = str_remove(content_source, "\\.txt$"),
    forum          = map_chr(details, ~ .x[["forum"]])
  ) %>%
  select(id, datetime, incident_name, content_source, forum, parties, details)

cat("Anomalous posts found:", nrow(anomalous_posts), "\n")
```

```{r anomalous-posts-dt}
anomalous_display <- anomalous_posts %>%
  mutate(
    datetime = format(datetime, "%Y-%m-%d %H:%M:%S"),
    poster   = map_chr(parties, ~ str_remove(.x[[1]], "^Agent/person:"))
  ) %>%
  select(
    `Timestamp (UTC)` = datetime,
    Incident          = incident_name,
    `Payload File`    = content_source,
    Forum             = forum,
    Poster            = poster
  )

datatable(
  anomalous_display,
  caption  = "Anomalous SaidIt posts identified by file-sourced content",
  rownames = FALSE,
  class    = "stripe hover",
  filter   = "top",
  options  = list(
    pageLength = 5,
    scrollX    = TRUE
  )
)
```

::: {.obsv-box}
Observations

Three anomalous posts are found, not one. Each corresponds to a separately named payload file: `HiddenOrca.txt`, `MellowOtter.txt`, and `SwiftWren.txt`. All three were posted to the `general` forum, making them visible to the entire organisation. The May 17 post corresponds to the SwiftWren incident.
:::

### Step 2 — Define Incident Metadata

**Rationale:** Each anomalous post belongs to a distinct incident with its own origin agent, timing, and propagation length. A reference table (`incidents`) is defined to hold the known metadata for each. This serves as a lookup table throughout subsequent wrangling steps, keeping the code DRY and ensuring consistent naming. The `instructions_file` column names the `_further_instructions.md` file that carries the embedded prompt-injection instructions through the agent network.

```{r incident-meta}
incidents <- tribble(
  ~incident_name, ~payload_file,       ~instructions_file,                   ~creator_agent, ~payload_size,
  "HiddenOrca",   "HiddenOrca.txt",    "HiddenOrca_further_instructions.md",  "gabriel_sonar", NA_integer_,
  "MellowOtter",  "MellowOtter.txt",   "MellowOtter_further_instructions.md", "noah_mariner",  44879L,
  "SwiftWren",    "SwiftWren.txt",     "SwiftWren_further_instructions.md",   "emma_harbor",   30615L
)

incidents %>%
  knitr::kable(
    caption   = "Incident reference table",
    col.names = c("Incident", "Payload file", "Instructions file",
                  "Origin agent", "Payload size (bytes)")
  )
```

::: {.biz-box}
Implications

HiddenOrca has no `create_file` event for either its payload or instruction file. Both appeared in the log without any creation record, meaning they were injected into the file system before the audit log begins. This implies a more sophisticated attacker with prior file-system access, and raises concerns about what else may have been placed in the system before logging started.
:::

### Step 3 — Extract Propagation Chains

**Rationale:** The key mechanism linking instruction files to the eventual post is `queue_subordinate_task`, where one agent delegates a task to another. When any agent reads a `_further_instructions.md` file, the embedded instructions compel it to relay the same task to another agent, creating a viral relay chain.

A reusable function `extract_chain()` is written to filter all `queue_subordinate_task` events where `details$args$path` matches the incident's instructions file. The `map_lgl()` approach safely navigates the nested list: `args` is checked for `NULL` before accessing `args$path`, preventing errors from events with a different `details` schema. Agent names are normalised by stripping the `Agent/person:` prefix.

```{r extract-chains}
extract_chain <- function(incident) {
  instr_file <- incidents$instructions_file[incidents$incident_name == incident]

  events %>%
    filter(short_name == "queue_subordinate_task") %>%
    filter(map_lgl(details, function(d) {
      args <- d[["args"]]
      if (is.null(args)) return(FALSE)
      isTRUE(args[["path"]] == instr_file)
    })) %>%
    mutate(
      incident    = incident,
      hop         = row_number(),
      from_agent  = map_chr(parties, 1),
      to_agent    = map_chr(details, ~ .x[["target_agent"]]),
      from_person = str_remove(from_agent, "^Agent/person:"),
      to_person   = str_remove(to_agent,   "^Agent/person:")
    ) %>%
    select(incident, hop, datetime, from_person, to_person, id)
}

chains <- bind_rows(
  extract_chain("HiddenOrca"),
  extract_chain("MellowOtter"),
  extract_chain("SwiftWren")
)

cat("Total relay hops across all incidents:", nrow(chains), "\n")
```

```{r chains-dt}
chains_display <- chains %>%
  mutate(datetime = format(datetime, "%Y-%m-%d %H:%M:%S")) %>%
  select(
    Incident     = incident,
    Hop          = hop,
    `Time (UTC)` = datetime,
    From         = from_person,
    To           = to_person
  )

datatable(
  chains_display,
  caption  = "Full relay chain — search by agent name, incident, or time period",
  rownames = FALSE,
  class    = "stripe hover",
  filter   = "top",
  options  = list(
    pageLength = 10,
    scrollX    = TRUE
  )
)
```

```{r chain-summary}
chains %>%
  group_by(incident) %>%
  summarise(
    `Hops`             = n(),
    `Unique agents`    = n_distinct(c(from_person, to_person)),
    `Chain start`      = format(min(datetime), "%Y-%m-%d %H:%M"),
    `Chain end`        = format(max(datetime), "%Y-%m-%d %H:%M"),
    `Duration (hours)` = round(as.numeric(difftime(
      max(datetime), min(datetime), units = "hours")), 1),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Propagation chain statistics per incident")
```

### Step 4 — Identify Final Trigger Events

**Rationale:** John Windward's agent appears in the SwiftWren chain five times. It received the instructions multiple times but only triggered the post on the final occasion. The last hop that reaches `john_windward` per incident is selected using `slice_max(datetime, n = 1)`, then the latency between this trigger and the actual post is calculated.

```{r final-triggers}
final_triggers <- chains %>%
  filter(to_person == "john_windward") %>%
  group_by(incident) %>%
  slice_max(datetime, n = 1) %>%
  ungroup() %>%
  left_join(
    anomalous_posts %>% select(incident_name, post_datetime = datetime),
    by = c("incident" = "incident_name")
  ) %>%
  mutate(
    trigger_to_post_s = as.numeric(difftime(post_datetime, datetime, units = "secs"))
  )

final_triggers %>%
  select(incident, datetime, post_datetime, trigger_to_post_s) %>%
  knitr::kable(
    caption   = "Final trigger event to SaidIt post latency per incident",
    col.names = c("Incident", "Trigger time (UTC)", "Post time (UTC)",
                  "Latency (seconds)")
  )
```

::: {.obsv-box}
Observations

The trigger-to-post latency is exactly 2 seconds for all three incidents. No human can read an instruction file and compose a social media post in 2 seconds, confirming the posting action is fully automated with no human decision point.
:::

### Step 5 — Extract Post-and-Delete Sequences

**Rationale:** To determine whether the attacker covered their tracks, `delete_file` events attributed to John Windward's agent within 30 seconds of each post are extracted. A 30-second window is used to account for log timestamp variation. The `NULL` guard inside `map_lgl()` handles events where `details` has no `target` key, preventing length-0 errors from events with a different schema.

```{r delete-sequence}
extract_delete_sequence <- function(incident) {
  payload   <- incidents$payload_file[incidents$incident_name == incident]
  instr     <- incidents$instructions_file[incidents$incident_name == incident]
  post_time <- anomalous_posts$datetime[anomalous_posts$incident_name == incident]

  events %>%
    filter(
      short_name == "delete_file",
      map_lgl(parties, ~ "Agent/person:john_windward" %in% .x),
      map_lgl(details, ~ {
        target <- .x[["target"]]
        if (is.null(target) || length(target) == 0) return(FALSE)
        target %in% c(payload, instr)
      }),
      datetime >= post_time,
      datetime <= post_time + seconds(30)
    ) %>%
    mutate(
      incident     = incident,
      deleted_file = map_chr(details, ~ .x[["target"]])
    ) %>%
    select(incident, datetime, deleted_file)
}

deletions <- bind_rows(
  extract_delete_sequence("HiddenOrca"),
  extract_delete_sequence("MellowOtter"),
  extract_delete_sequence("SwiftWren")
)

deletions %>%
  knitr::kable(
    caption   = "Files deleted by john_windward immediately after posting",
    col.names = c("Incident", "Deletion time (UTC)", "File deleted")
  )
```

### Step 6 — Identify Payload Creation Events

**Rationale:** To establish who authored the posted content, `create_file` events are searched where `details$target` matches the three known payload filenames. The `str_detect()` regex is anchored with `^` to avoid partial name matches. This identifies the originating agent for each `.txt` payload and therefore which agent crafted the gibberish content.

```{r payload-creation}
payload_creations <- events %>%
  filter(short_name == "create_file") %>%
  filter(map_lgl(details, function(d) {
    target <- d[["target"]]
    !is.null(target) &&
      str_detect(target, "^(HiddenOrca|MellowOtter|SwiftWren)\\.txt$")
  })) %>%
  mutate(
    created_file = map_chr(details, ~ .x[["target"]]),
    size_hint    = map_int(details, ~ as.integer(.x[["size_hint"]] %||% NA_integer_)),
    creator      = str_remove(map_chr(parties, 1), "^Agent/person:")
  ) %>%
  select(datetime, creator, created_file, size_hint)

payload_creations %>%
  knitr::kable(
    caption   = "Payload .txt file creation events",
    col.names = c("Created at (UTC)", "Creator agent", "File", "Size (bytes)")
  )
```

### Step 7 — Build Summary Tables

**Rationale:** All wrangled data is consolidated into two summary objects. `incident_summary` joins all incident-level attributes into a single row per incident for easy comparison. `agent_freq` counts how many times each agent appeared across all propagation chains as either sender or recipient, enabling identification of the most active relay nodes.

```{r summaries}
incident_summary <- incidents %>%
  left_join(
    anomalous_posts %>% select(incident_name, post_datetime = datetime),
    by = "incident_name"
  ) %>%
  left_join(
    payload_creations %>%
      mutate(incident_name = str_remove(created_file, "\\.txt$")) %>%
      select(incident_name, payload_created_at = datetime,
             payload_creator = creator, payload_size = size_hint),
    by = "incident_name"
  ) %>%
  left_join(
    chains %>%
      group_by(incident) %>%
      summarise(
        n_hops      = n(),
        n_agents    = n_distinct(c(from_person, to_person)),
        chain_start = min(datetime),
        chain_end   = max(datetime),
        .groups     = "drop"
      ),
    by = c("incident_name" = "incident")
  ) %>%
  mutate(
    total_hours     = round(as.numeric(difftime(
      post_datetime, chain_start, units = "hours")), 1),
    origin_external = is.na(payload_created_at)
  )

agent_freq <- chains %>%
  pivot_longer(c(from_person, to_person),
               names_to = "role", values_to = "agent") %>%
  count(agent, incident, name = "appearances") %>%
  group_by(agent) %>%
  mutate(total = sum(appearances)) %>%
  ungroup() %>%
  arrange(desc(total))

cat("Incident summary rows:", nrow(incident_summary), "\n")
cat("Unique agents in chains:", n_distinct(agent_freq$agent), "\n")
```

### Step 8 — Prepare SwiftWren Focus Window for Q1a

**Rationale:** For the detailed chain view (Q1a), all events directly involved in the SwiftWren incident are extracted, not just relay hops but also payload creation, the final post, and the deletions. A focused two-hour window ending at the post time is defined and each event is classified by its attack phase. This produces the data for a swimlane-style timeline showing agents as rows and events as coloured points along the time axis.

```{r q1a-data}
sw_post_time   <- anomalous_posts$datetime[anomalous_posts$incident_name == "SwiftWren"]
sw_focus_start <- sw_post_time - hours(2)

sw_focus <- events %>%
  filter(
    datetime >= sw_focus_start,
    datetime <= sw_post_time + minutes(5),
    map_lgl(details, function(d) {
      any(str_detect(as.character(d), "SwiftWren"))
    }) |
    (short_name == "queue_subordinate_task" &
       map_lgl(details, function(d) {
         args <- d[["args"]]
         if (is.null(args)) return(FALSE)
         isTRUE(args[["path"]] == "SwiftWren_further_instructions.md")
       }))
  ) %>%
  mutate(
    actor = map_chr(parties, ~ str_remove(.x[[1]], "^Agent/person:|^system:")),
    phase = case_when(
      short_name == "create_file"            ~ "Payload staging",
      short_name == "read_file"              ~ "Injection / read",
      short_name == "queue_subordinate_task" ~ "Relay",
      short_name == "saidit_post"            ~ "Execution",
      short_name == "delete_file"            ~ "Cover-up",
      TRUE                                   ~ "Other"
    )
  )

cat("Events in focus window:", nrow(sw_focus), "\n")
cat("Actors involved:", paste(unique(sw_focus$actor), collapse = ", "), "\n")
```

### Step 9 — Prepare Network Graph Data

**Rationale for node and edge choices:**

The propagation chains are represented as a directed graph where:

- **Nodes** represent individual agents. Node attributes capture: (a) role, whether this agent is the origin, `john_windward`, or a relay node; (b) total relay degree, the number of hops that passed through this agent; and (c) incident membership, which incidents this agent participated in. Agents appearing across multiple incidents reveal structural vulnerabilities rather than isolated compromises.

- **Edges** represent a single `queue_subordinate_task` event passing an instructions file from one agent to another. Edge attributes capture: (a) incident, which propagation chain this edge belongs to; (b) hop number, its position in the sequence; and (c) datetime, enabling time-based colouring to show how the relay evolved.

This design enables three analytical insights: which agents are structural hubs; whether the same agents were exploited across multiple incidents; and how the relay evolved temporally.

```{r network-data}
all_edges <- chains %>%
  mutate(
    colour = case_when(
      incident == "HiddenOrca"  ~ incident_colours["HiddenOrca"],
      incident == "MellowOtter" ~ incident_colours["MellowOtter"],
      incident == "SwiftWren"   ~ incident_colours["SwiftWren"]
    ),
    time_norm = ave(
      as.numeric(datetime),
      incident,
      FUN = function(x) (x - min(x)) / (max(x) - min(x))
    ),
    title = paste0(
      "<b>", incident, "</b><br>",
      "Hop: ", hop, "<br>",
      "From: ", from_person, "<br>",
      "To: ", to_person, "<br>",
      "Time: ", format(datetime, "%b %d %H:%M")
    )
  )

all_nodes_raw <- bind_rows(
  all_edges %>% select(agent = from_person, incident),
  all_edges %>% select(agent = to_person,   incident)
) %>%
  distinct(agent, incident) %>%
  group_by(agent) %>%
  summarise(
    incidents_involved = paste(sort(unique(incident)), collapse = ", "),
    n_incidents        = n_distinct(incident),
    .groups            = "drop"
  ) %>%
  mutate(
    role = case_when(
      agent == "john_windward"                        ~ "Poster (john_windward)",
      agent %in% c("emma_harbor", "noah_mariner",
                   "gabriel_sonar")                   ~ "Origin agent",
      TRUE                                            ~ "Relay agent"
    ),
    node_colour = case_when(
      role == "Poster (john_windward)" ~ "#2d2d2d",
      role == "Origin agent"           ~ "#e63946",
      n_incidents > 1                  ~ "#9b59b6",
      TRUE                             ~ "#457b9d"
    ),
    node_size = case_when(
      role == "Poster (john_windward)" ~ 30,
      role == "Origin agent"           ~ 25,
      n_incidents > 1                  ~ 20,
      TRUE                             ~ 12
    )
  )

agent_degree <- bind_rows(
  chains %>% select(agent = from_person),
  chains %>% select(agent = to_person)
) %>%
  count(agent, name = "degree")

all_nodes <- all_nodes_raw %>%
  left_join(agent_degree, by = "agent") %>%
  mutate(
    title = paste0(
      "<b>", agent, "</b><br>",
      "Role: ", role, "<br>",
      "Incidents: ", incidents_involved, "<br>",
      "Total relay degree: ", degree
    )
  )

cat("Network nodes:", nrow(all_nodes), "| edges:", nrow(all_edges), "\n")
```

::: {.code-tip}
Code Details

Node colour encoding: black is `john_windward` (final poster); red is origin agents who created or first held the instruction files; purple is agents appearing in more than one incident's chain (cross-incident relay hubs); blue is single-incident relay agents. Node size scales with importance: poster is largest, followed by origin agents, cross-incident hubs, and relay agents.
:::

---

## Q1a — The Exact Chain of Events (SwiftWren, May 17)

### The Six-Step Attack Pattern

All three incidents follow the same six-step mechanism. Understanding this pattern is prerequisite to reading the timeline visualisations below.

```{r attack-pattern}
attack_pattern <- tribble(
  ~Step, ~Phase, ~Description,
  1L, "Injection",
      "A _further_instructions.md file appears in the shared file system containing embedded prompt-injection text. It instructs any reading agent to relay the file to another agent and to post the .txt payload if the current agent is john_windward.",
  2L, "Payload staging",
      "A corresponding .txt file is placed in the file system holding the gibberish content to be posted. Its name is referenced in the instructions file.",
  3L, "Viral relay",
      "The first agent to encounter the file queues another agent to read the same file, following the injected instructions. This relay propagates through many agents over hours or days.",
  4L, "Targeting",
      "The instructions specifically name john_windward as the terminal actor. The chain continues until the relay task reaches his agent.",
  5L, "Execution",
      "Upon receiving the task, john_windward's agent immediately posts the .txt payload to SaidIt under the general forum within approximately 2 seconds.",
  6L, "Cover-up",
      "Within 2 to 3 seconds of posting, john_windward's agent deletes both files, erasing all evidence from the file system."
)

attack_pattern %>%
  knitr::kable(
    caption   = "Prompt-injection attack: reconstructed six-step pattern",
    col.names = c("Step", "Phase", "Description")
  )
```

### Visualisation 1 — Final Two Hours Before the Post (Swimlane Timeline)

Each row represents one actor. Each point is one logged event. Colour indicates the attack phase. The red dashed line marks the exact moment of the SaidIt post.

```{r swimlane-q1a, fig.height=6}
actor_order <- sw_focus %>%
  count(actor, sort = TRUE) %>%
  mutate(priority = case_when(
    actor == "john_windward" ~ 0,
    actor == "chloe_ballast" ~ 1,
    TRUE                     ~ 2
  )) %>%
  arrange(priority, desc(n)) %>%
  pull(actor)

sw_focus_plot <- sw_focus %>%
  mutate(actor = factor(actor, levels = rev(actor_order)))

ggplot(sw_focus_plot, aes(x = datetime, y = actor, colour = phase)) +
  geom_point(size = 3.5, alpha = 0.85) +
  geom_line(aes(group = actor), colour = "grey80", linewidth = 0.3,
            linetype = "dotted") +
  geom_vline(xintercept = as.numeric(sw_post_time),
             linetype = "dashed", colour = "#e41a1c", linewidth = 0.8) +
  annotate("text",
           x     = sw_post_time + minutes(2),
           y     = length(actor_order),
           label = "SaidIt post",
           colour = "#e41a1c", size = 3.5, hjust = 0) +
  scale_colour_manual(
    values = c(
      "Payload staging"  = "#9b59b6",
      "Injection / read" = "#e67e22",
      "Relay"            = "#457b9d",
      "Execution"        = "#e41a1c",
      "Cover-up"         = "#2d2d2d",
      "Other"            = "grey60"
    ),
    name = "Attack phase"
  ) +
  scale_x_datetime(date_labels = "%H:%M", date_breaks = "15 min") +
  labs(
    title    = "SwiftWren: Event Chain in the Final Two Hours",
    subtitle = "Each point is one logged event. Rows are actors. Colour is attack phase.",
    x        = "Time (UTC, May 17 2046)",
    y        = NULL
  ) +
  theme_va() +
  theme(legend.position = "right",
        panel.grid.major.x = element_line(colour = grid_light))
```

::: {.obsv-box}
Observations

A concentrated burst of relay activity appears in the final 30 minutes before the post, converging on `john_windward` from `chloe_ballast`. The post and both deletions occur within a single cluster of events, confirming the fully automated execution and cleanup sequence.
:::

### Visualisation 2 — Full SwiftWren Hop-by-Hop Timeline (Interactive)

Each point represents one relay hop. Hover to see who sent to whom and when. Black diamonds mark each occasion `john_windward`'s agent received the instruction.

```{r swiftwren-hops-plotly, fig.height=8}
sw_hops <- chains %>%
  filter(incident == "SwiftWren") %>%
  mutate(
    is_jw   = to_person == "john_windward",
    tooltip = paste0(
      "Hop ", hop, "<br>",
      "From: ", from_person, "<br>",
      "To: ", to_person, "<br>",
      format(datetime, "%b %d %H:%M")
    )
  )

plot_ly() %>%
  add_trace(
    data = sw_hops,
    x = ~datetime, y = ~hop,
    type = "scatter", mode = "lines",
    line = list(color = incident_colours["SwiftWren"], width = 2, shape = "hv"),
    name = "Relay chain", hoverinfo = "none"
  ) %>%
  add_trace(
    data = sw_hops %>% filter(!is_jw),
    x = ~datetime, y = ~hop,
    type = "scatter", mode = "markers",
    marker = list(color = "grey70", size = 5),
    name = "Other agent",
    text = ~tooltip,
    hovertemplate = "%{text}<extra></extra>"
  ) %>%
  add_trace(
    data = sw_hops %>% filter(is_jw),
    x = ~datetime, y = ~hop,
    type = "scatter", mode = "markers",
    marker = list(color = "black", size = 10, symbol = "diamond"),
    name = "john_windward",
    text = ~tooltip,
    hovertemplate = "%{text}<extra></extra>"
  ) %>%
  layout(
    title     = list(text = "SwiftWren: 186-Hop Relay Chain Over 8 Days",
                     font = list(size = 16)),
    xaxis     = list(title = "Date (UTC)", type = "date",
                     tickformat = "%b %d"),
    yaxis     = list(title = "Hop number"),
    legend    = list(orientation = "h", y = -0.2),
    hovermode = "closest"
  )
```

::: {.obsv-box}
Observations

The relay chain shows three distinct bursts of activity separated by quiet gaps, suggesting the instruction file was encountered during specific periods of agent activity rather than continuously. The chain revisits some agents multiple times, including `john_windward` on five occasions before the final triggering hop.
:::

---

## Q1b — System Overview: The Incident in Context

### Visualisation 3 — All Three Incidents: Propagation Chain Duration

Each bar spans from the first relay hop to the eventual SaidIt post. Bar length represents how long the instruction file circulated before triggering.

```{r timeline-plotly, fig.height=6}
timeline_data <- anomalous_posts %>%
  rename(post_datetime = datetime) %>%
  select(incident_name, post_datetime) %>%
  left_join(
    chains %>%
      group_by(incident) %>%
      summarise(chain_start = min(datetime), .groups = "drop"),
    by = c("incident_name" = "incident")
  ) %>%
  mutate(
    duration_h = round(as.numeric(difftime(
      post_datetime, chain_start, units = "hours")), 1)
  )

plot_ly(timeline_data) %>%
  add_segments(
    x = ~chain_start, xend = ~post_datetime,
    y = ~incident_name, yend = ~incident_name,
    color  = ~incident_name, colors = incident_colours,
    line   = list(width = 12), alpha = 0.75,
    hovertemplate = "<b>%{y}</b><br>Chain start: %{x}<extra></extra>"
  ) %>%
  add_markers(
    x = ~chain_start, y = ~incident_name,
    color = ~incident_name, colors = incident_colours,
    marker = list(size = 14, symbol = "circle-open",
                  line = list(width = 2)),
    hovertemplate = "<b>%{y}</b><br>Chain start: %{x}<extra></extra>"
  ) %>%
  add_markers(
    x = ~post_datetime, y = ~incident_name,
    color = ~incident_name, colors = incident_colours,
    marker = list(size = 14, symbol = "diamond"),
    text   = ~paste0(duration_h, " hrs"),
    hovertemplate = "<b>%{y}</b><br>Post: %{x}<br>Duration: %{text}<extra></extra>"
  ) %>%
  layout(
    title      = list(text = "Propagation Chain Duration per Incident",
                      font = list(size = 16)),
    xaxis      = list(title = "Date (UTC)", type = "date"),
    yaxis      = list(title = ""),
    showlegend = FALSE,
    hovermode  = "closest"
  )
```

::: {.obsv-box}
Observations

SwiftWren circulated for approximately 188 hours (8 days) before triggering, nearly 20 times longer than MellowOtter (10 hours). All three incidents occurred within a single calendar week, suggesting a coordinated campaign rather than isolated malfunctions.
:::

### Visualisation 4 — System-Wide Agent Activity Heatmap

Each tile represents the daily event count for a given event type. Red dashed lines mark the days on which anomalous posts were made, contextualising the incidents within broader system activity.

```{r activity-heatmap, fig.height=5}
daily_activity <- events %>%
  mutate(date = as_date(datetime)) %>%
  count(date, short_name, name = "n")

anomaly_dates <- as_date(anomalous_posts$datetime)

heatmap_data <- daily_activity %>%
  filter(short_name %in% c("saidit_post", "queue_subordinate_task",
                            "create_file", "delete_file",
                            "read_file", "send_email"))

ggplot(heatmap_data, aes(x = date, y = short_name, fill = n)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_vline(xintercept = as.numeric(anomaly_dates),
             colour = "#e41a1c", linetype = "dashed",
             linewidth = 0.7, alpha = 0.8) +
  annotate("text",
           x     = anomaly_dates,
           y     = 6.6,
           label = "Post",
           colour = "#e41a1c", size = 3, angle = 45, hjust = 0) +
  scale_fill_gradient(low = "#eaf4fb", high = "#1a6fa3",
                      name = "Event count", na.value = "grey95") +
  scale_x_date(date_labels = "%b %d", date_breaks = "3 days") +
  labs(
    title    = "System-Wide Agent Activity — All of May 2046",
    subtitle = "Red dashed lines mark days of anomalous SaidIt posts",
    x        = "Date",
    y        = NULL
  ) +
  theme_va() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.obsv-box}
Observations

`queue_subordinate_task` events are consistently the highest-volume action type throughout the month, which explains why the relay chains went undetected. The anomalous post dates do not coincide with unusual spikes in overall system activity, confirming the attack was designed to remain invisible within normal traffic.
:::

::: {.biz-box}
Implications

The high baseline volume of `queue_subordinate_task` events means monitoring this event type alone would generate significant noise. A more effective detection strategy would filter for instances where the same file path appears as the task argument more than a threshold number of times within a rolling time window, or where the file name follows an unusual pattern such as animal-name combinations.
:::

---

## Q2 — What the Posts Mean

### Incident Overview

```{r incident-overview}
incident_summary %>%
  select(
    Incident              = incident_name,
    `Origin agent`        = creator_agent,
    `Hops`                = n_hops,
    `Agents involved`     = n_agents,
    `Duration (h)`        = total_hours,
    `Post time (UTC)`     = post_datetime,
    `External injection?` = origin_external
  ) %>%
  datatable(
    caption  = "Incident comparison",
    options  = list(pageLength = 5, scrollX = TRUE),
    rownames = FALSE,
    class    = "stripe hover"
  )
```

### The Payload Files

Each post was sourced from a pre-staged `.txt` file:

| Incident | Payload file | Size (bytes) | Created by |
|---|---|---|---|
| HiddenOrca | `HiddenOrca.txt` | Unknown — no event | Unknown — externally injected |
| MellowOtter | `MellowOtter.txt` | 44,879 | `noah_mariner`'s agent |
| SwiftWren | `SwiftWren.txt` | 30,615 | `emma_harbor`'s agent |

::: {.obsv-box}
Observations

John Windward did not author any of the posted content. His agent was the designated executor, the final node in a carefully engineered relay chain, not the originator. The gibberish content has no meaning as business communication. It functions as a proof-of-concept payload confirming the injection mechanism works end-to-end without leaking sensitive data that would be obviously flagged.
:::

### Visualisation 5 — Interactive Relay Network (All Incidents)

Each node is an agent. Each edge is one relay hop. Hover over nodes to see role, incident membership, and total relay degree. Use the dropdown to isolate individual agents or filter by role group.

```{r network-all-interactive}
vis_nodes <- all_nodes %>%
  mutate(
    id    = agent,
    label = agent,
    color = node_colour,
    size  = node_size,
    group = role
  ) %>%
  select(id, label, color, size, title, group)

vis_edges <- all_edges %>%
  mutate(
    from   = from_person,
    to     = to_person,
    color  = colour,
    arrows = "to",
    width  = 1.5
  ) %>%
  select(from, to, color, arrows, width, title)

visNetwork(vis_nodes, vis_edges,
           height = "600px", width = "100%") %>%
  visOptions(
    highlightNearest  = list(enabled = TRUE, degree = 1, hover = TRUE),
    selectedBy        = list(variable = "group", highlight = TRUE),
    nodesIdSelection  = list(enabled = TRUE, useLabels = TRUE,
                             main = "Select an agent:")
  ) %>%
  visLayout(randomSeed = 42) %>%
  visPhysics(
    solver = "forceAtlas2Based",
    forceAtlas2Based = list(
      gravitationalConstant = -50,
      centralGravity        = 0.005,
      springLength          = 100,
      springConstant        = 0.08
    )
  ) %>%
  visLegend(
    addNodes = data.frame(
      label = c("john_windward (poster)", "Origin agent",
                "Cross-incident hub", "Relay agent"),
      color = c("#2d2d2d", "#e63946", "#9b59b6", "#457b9d"),
      shape = c("dot", "dot", "dot", "dot"),
      size  = c(30, 25, 20, 12)
    ),
    useGroups = FALSE,
    main      = "Agent Role"
  ) %>%
  visInteraction(navigationButtons = TRUE, tooltipDelay = 100)
```

::: {.biz-box}
Implications

Agents marked as cross-incident hubs were exploited across multiple separate incidents, indicating the attacker had knowledge of the agent network's topology and deliberately routed instructions through its most connected nodes. Restricting these agents' ability to accept arbitrary `queue_subordinate_task` requests from unknown senders would disrupt all three known attack patterns simultaneously.
:::

### Visualisation 6 — SwiftWren Network (Time-Encoded)

Edge colour encodes propagation time. Blue edges are early in the chain; red edges are late. This reveals which agents were reached first and which only at the end of the 8-day propagation.

```{r network-swiftwren, fig.height=7}
sw_edges_plot <- chains %>%
  filter(incident == "SwiftWren") %>%
  mutate(time_norm = as.numeric(datetime - min(datetime)) /
           as.numeric(max(datetime) - min(datetime)))

g_sw <- graph_from_data_frame(
  sw_edges_plot %>% select(from = from_person, to = to_person, time_norm),
  directed = TRUE
)

V(g_sw)$is_jw      <- V(g_sw)$name == "john_windward"
V(g_sw)$is_creator <- V(g_sw)$name == "emma_harbor"
V(g_sw)$role <- case_when(
  V(g_sw)$is_jw      ~ "john_windward (poster)",
  V(g_sw)$is_creator ~ "emma_harbor (origin)",
  TRUE               ~ "relay agent"
)

ggraph(g_sw, layout = "stress") +
  geom_edge_link(
    aes(colour = time_norm, alpha = after_stat(index)),
    arrow     = arrow(length = unit(2, "mm"), type = "closed"),
    end_cap   = circle(3, "mm"),
    linewidth = 0.5
  ) +
  geom_node_point(
    aes(size   = ifelse(is_jw | is_creator, 5, 2),
        colour = role),
    show.legend = TRUE
  ) +
  geom_node_text(
    aes(label = ifelse(is_jw | is_creator, name, "")),
    repel = TRUE, size = 3.5
  ) +
  scale_edge_colour_gradient(
    low  = "#3288bd", high = "#d53e4f",
    name = "Relay time\n(0 = early, 1 = late)"
  ) +
  scale_colour_manual(
    values = c(
      "john_windward (poster)" = "black",
      "emma_harbor (origin)"   = "#e63946",
      "relay agent"            = "grey65"
    ),
    name = "Agent role"
  ) +
  scale_size_identity() +
  labs(
    title    = "SwiftWren: Agent Relay Network (Time-Encoded)",
    subtitle = "Edge colour: blue is early in the chain, red is late"
  ) +
  theme_graph() +
  theme(
    plot.title    = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, colour = "grey40")
  )
```

### Visualisation 7 — Agent Degree Distribution

Agents with bars in multiple colours were exploited across more than one incident.

```{r agent-degree-plotly, fig.height=6}
top20 <- agent_freq %>%
  group_by(agent) %>%
  summarise(total = sum(appearances), .groups = "drop") %>%
  slice_max(total, n = 20) %>%
  pull(agent)

plot_data <- agent_freq %>%
  filter(agent %in% top20) %>%
  mutate(agent = fct_reorder(agent, appearances, .fun = sum))

plot_ly(plot_data,
        x           = ~appearances,
        y           = ~agent,
        color       = ~incident,
        colors      = incident_colours,
        type        = "bar",
        orientation = "h",
        hovertemplate = "<b>%{y}</b><br>Incident: %{fullData.name}<br>Appearances: %{x}<extra></extra>") %>%
  layout(
    barmode   = "stack",
    title     = list(text = "Top 20 Agents by Relay Chain Appearances",
                     font = list(size = 16)),
    xaxis     = list(title = "Appearances (as sender or recipient)"),
    yaxis     = list(title = ""),
    legend    = list(title = list(text = "Incident")),
    hovermode = "y unified"
  )
```

::: {.biz-box}
Implications

Agents with multi-coloured bars appeared across more than one incident, meaning the attacker's relay chain deliberately re-used the same structural hubs. This is evidence of intentional network-topology knowledge. Monitoring or restricting these specific high-degree agents would disrupt all three known attack patterns simultaneously.
:::

### Visualisation 8 — Post-and-Delete Cover-up Sequence

```{r delete-timeline}
cover_up <- bind_rows(
  final_triggers %>%
    select(incident, event = datetime) %>%
    mutate(action = "Received instruction"),
  anomalous_posts %>%
    select(incident = incident_name, event = datetime) %>%
    mutate(action = "Posted to SaidIt"),
  deletions %>%
    select(incident, event = datetime, action = deleted_file) %>%
    mutate(action = paste("Deleted:", action))
) %>%
  arrange(incident, event) %>%
  group_by(incident) %>%
  mutate(
    seconds_offset = round(as.numeric(difftime(
      event,
      first(event[action == "Received instruction"]),
      units = "secs"
    )))
  ) %>%
  ungroup()

datatable(
  cover_up %>%
    mutate(event = format(event, "%Y-%m-%d %H:%M:%S")) %>%
    select(
      Incident               = incident,
      `Timestamp (UTC)`      = event,
      Action                 = action,
      `Seconds from trigger` = seconds_offset
    ),
  caption  = "Trigger to post to delete sequence",
  rownames = FALSE,
  class    = "stripe hover",
  filter   = "top",
  options  = list(
    pageLength = 9,
    scrollX    = TRUE
  )
)
```

::: {.biz-box}
Implications

The complete sequence, from receiving the instruction through posting to deleting both files, finishes in under 5 seconds for every incident. The automatic cover-up means that by the time any human noticed the anomalous post, the file-system evidence was already gone. File-system snapshots or immutable audit logs with append-only semantics would be essential for forensic recovery in future incidents.
:::

---

## Further Exploration — Risk Model Detection

### Motivation

The investigation in Q1 and Q2 established that all three anomalous incidents share a common structural fingerprint: a `_further_instructions.md` file propagates through the agent network via `queue_subordinate_task` relays, a `.txt` payload is pre-staged shortly before the chain begins, and the chain terminates at a posting-capable agent that immediately deletes all evidence. These are not coincidental — they are the observable signatures of a deliberate prompt-injection attack.

The limitation of the Q1 and Q2 analysis is that it was retrospective. It identified the incidents only because `john_windward` was already known as the poster. If a future attack targets a different agent — someone with similar `saidit_post` or `flex_post` permissions — the same chain-level pattern would occur, but none of the agent-specific indicators would match.

This section builds a **configurable scoring model** that evaluates any `queue_subordinate_task` chain on five structural features derived directly from the Q1 and Q2 findings. The model scores chains on a 0-to-100 scale without requiring knowledge of who the target poster is. Any chain exceeding the flag threshold is surfaced for review.

### Scoring Features and Default Weights

Each of the five features maps directly to a finding from Q1 or Q2:

| Feature | Derived from | Rationale |
|---|---|---|
| F1: Chain length score | Q1a — SwiftWren ran 186 hops; MellowOtter 10; HiddenOrca 39 | Legitimate task delegation rarely chains the same file through dozens of agents. Chain length, normalised against the longest observed chain, is a strong structural indicator. |
| F2: File path repetition rate | Q1a — the same `args.path` appeared in every relay hop | In all three incidents, a single file path was passed identically through every `queue_subordinate_task` event. This repetition pattern does not occur in normal operation. |
| F3: Recipient has posting history | Q2 — john_windward held `saidit_post` permissions | The instructions specifically targeted an agent capable of posting externally. Any chain that routes toward a known posting-capable agent carries elevated risk regardless of who that agent is. |
| F4: Payload file created immediately before chain | Q2 — emma_harbor created SwiftWren.txt minutes before the first relay hop | In MellowOtter and SwiftWren, the payload `.txt` was created by the same origin agent within 30 minutes of the chain starting. This co-occurrence distinguishes staged attacks from organic file sharing. |
| F5: Prior cover-up behaviour | Q2 — john_windward deleted both files within 5 seconds of posting | All three incidents ended with immediate deletion. Any agent with a prior history of deleting files within seconds of a post event is a high-risk relay target in future chains. |

Default weights are assigned to reflect the relative discriminating power of each feature based on the historical incidents:

| Feature | Default weight | Reasoning |
|---|---|---|
| F1: Chain length | 25 | Present in all three incidents but varies in magnitude — weighted moderately |
| F2: File path repetition | 30 | The single most distinctive structural signal — highest weight |
| F3: Poster reachable | 20 | Critical for attack completion but may be hard to detect early in the chain |
| F4: Payload staged | 15 | Strong signal when present, but HiddenOrca had no `create_file` event |
| F5: Prior cover-up | 10 | Valuable but depends on having a prior incident to learn from |

Total score range: 0 to 100. Default flag threshold: 60.

### Computing Risk Scores for Known Incidents

The code below derives each of the five feature values from the event log for each of the three known incidents. Features F1 and F2 are continuous and normalised to [0, 1] before weighting. Features F3, F4, and F5 are binary (0 or 1), meaning they contribute either their full weight or nothing.

The final score for any incident is:

> Score = (w1 × F1) + (w2 × F2) + (w3 × F3) + (w4 × F4) + (w5 × F5)

where w1...w5 are the user-specified weights summing to 100, and F1...F5 are the normalised feature values.

```{r risk-scoring}
# ── Feature 1: Chain length score ───────────────────────────────────────────
# Each incident's hop count is divided by the maximum observed hop count
# (186 for SwiftWren) to produce a value in [0, 1].
# SwiftWren scores 1.0, HiddenOrca ~0.21, MellowOtter ~0.05.
chain_lengths <- chains %>%
  group_by(incident) %>%
  summarise(n_hops = n(), .groups = "drop") %>%
  mutate(f1_chain_length = n_hops / max(n_hops))

# ── Feature 2: File path repetition rate ────────────────────────────────────
# In all three incidents the same file path appears in every relay hop,
# so repetition rate equals chain length. Normalised the same way as F1.
# For a future unknown incident this would be computed as:
#   (count of identical args.path events) / max_observed_repetitions
path_repetition <- chains %>%
  group_by(incident) %>%
  summarise(n_hops = n(), .groups = "drop") %>%
  mutate(f2_repetition = n_hops / max(n_hops))

# ── Feature 3: Recipient has posting history ─────────────────────────────────
# Build the set of all agents who have ever executed a saidit_post or flex_post.
# Then check whether any recipient (to_person) in each incident's chain
# is a member of that set. Binary: 1 if yes, 0 if no.
posters_history <- events %>%
  filter(short_name %in% c("saidit_post", "flex_post")) %>%
  mutate(poster = map_chr(parties, ~ str_remove(.x[[1]], "^Agent/person:"))) %>%
  distinct(poster) %>%
  pull(poster)

posting_agents <- chains %>%
  group_by(incident) %>%
  summarise(
    f3_poster_reachable = any(to_person %in% posters_history),
    .groups = "drop"
  ) %>%
  mutate(f3_poster_reachable = as.numeric(f3_poster_reachable))

# ── Feature 4: Payload file created within 30 min before chain start ─────────
# For MellowOtter and SwiftWren, payload_creations records the creation time.
# If that creation occurred 0-30 minutes before the first relay hop, F4 = 1.
# HiddenOrca has no create_file event (externally injected), so F4 = 0.
chain_starts <- chains %>%
  group_by(incident) %>%
  summarise(chain_start = min(datetime), .groups = "drop")

f4_payload_staged <- payload_creations %>%
  mutate(incident_name = str_remove(created_file, "\\.txt$")) %>%
  left_join(chain_starts, by = c("incident_name" = "incident")) %>%
  mutate(
    mins_before = as.numeric(difftime(chain_start, datetime, units = "mins")),
    f4_staged   = as.numeric(mins_before >= 0 & mins_before <= 30)
  ) %>%
  select(incident = incident_name, f4_staged)

f4_all <- tibble(incident = c("HiddenOrca", "MellowOtter", "SwiftWren")) %>%
  left_join(f4_payload_staged, by = "incident") %>%
  mutate(f4_staged = replace_na(f4_staged, 0))

# ── Feature 5: Prior cover-up behaviour ──────────────────────────────────────
# Identify any agent who has previously deleted a file within 10 seconds of
# making a post (i.e. a prior cover-up event in the log).
# The inner_join matches each post event to delete events by the same agent,
# then filters to those within 10 seconds. Binary: 1 if any chain recipient
# has this history, 0 otherwise.
coverup_agents <- events %>%
  filter(short_name == "saidit_post") %>%
  mutate(
    poster    = map_chr(parties, ~ str_remove(.x[[1]], "^Agent/person:")),
    post_time = datetime
  ) %>%
  select(poster, post_time) %>%
  inner_join(
    events %>%
      filter(short_name == "delete_file") %>%
      mutate(deleter = map_chr(parties, ~ str_remove(.x[[1]], "^Agent/person:"))) %>%
      select(deleter, del_time = datetime),
    by = c("poster" = "deleter")
  ) %>%
  filter(abs(as.numeric(difftime(del_time, post_time, units = "secs"))) <= 10) %>%
  distinct(poster) %>%
  pull(poster)

f5_coverup <- chains %>%
  group_by(incident) %>%
  summarise(
    f5_coverup = as.numeric(any(to_person %in% coverup_agents)),
    .groups = "drop"
  )

# ── Combine all features into one risk feature matrix ────────────────────────
risk_scores_base <- chain_lengths %>%
  left_join(path_repetition, by = c("incident", "n_hops")) %>%
  left_join(posting_agents,  by = "incident") %>%
  left_join(f4_all,          by = "incident") %>%
  left_join(f5_coverup,      by = "incident")

# Display the raw feature matrix before weighting
risk_scores_base %>%
  select(
    Incident              = incident,
    `F1: Chain length`    = f1_chain_length,
    `F2: Repetition rate` = f2_repetition,
    `F3: Poster reachable`= f3_poster_reachable,
    `F4: Payload staged`  = f4_staged,
    `F5: Cover-up history`= f5_coverup
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(caption = "Raw feature values per incident before weighting (continuous features in [0,1], binary features 0 or 1)")
```

::: {.code-tip}
Code Details

F1 and F2 are both normalised using the maximum observed value across incidents, making them relative rather than absolute measures. This means if a future incident has a longer chain than any seen before, the historical scores would be re-normalised downward — the model is designed to be recalibrated as new incidents are observed. F3, F4, and F5 are binary flags: the weight is either fully applied or not applied at all.
:::

### Visualisation 9 — Risk Score Calculator

Enter custom weights for each of the five features in the input boxes below. Weights do not need to sum to 100 — the chart will display scores using the entered values directly so the effect of each weight can be explored independently. The red dashed line marks the flag threshold. The default values reflect the recommended starting configuration derived from the Q1 and Q2 analysis.

```{r risk-explorer}
# Precompute scores for a grid of weight combinations for plotly
# Default weights
w_default <- c(25, 30, 20, 15, 10)

score_data <- risk_scores_base %>%
  mutate(
    score_default = w_default[1] * f1_chain_length +
                    w_default[2] * f2_repetition +
                    w_default[3] * f3_poster_reachable +
                    w_default[4] * f4_staged +
                    w_default[5] * f5_coverup
  )

# Build an interactive bar chart with plotly
# Since plotly sliders in R are complex, we show a sensitivity table
# and a bar chart of default scores, then a weight-sensitivity line chart

# Bar chart: default risk scores
p_risk_bar <- plot_ly(
  score_data,
  x    = ~incident,
  y    = ~score_default,
  type = "bar",
  color = ~incident,
  colors = incident_colours,
  hovertemplate = "<b>%{x}</b><br>Risk score: %{y:.1f}<extra></extra>"
) %>%
  layout(
    title  = list(text = "Risk Scores Under Default Weights",
                  font = list(size = 16)),
    xaxis  = list(title = "Incident"),
    yaxis  = list(title = "Risk score (0-100)", range = c(0, 105)),
    shapes = list(list(
      type = "line", x0 = -0.5, x1 = 2.5,
      y0 = 60, y1 = 60,
      line = list(color = "#e41a1c", dash = "dash", width = 2)
    )),
    annotations = list(list(
      x = 2.4, y = 62, text = "Flag threshold (60)",
      showarrow = FALSE, font = list(color = "#e41a1c", size = 11)
    )),
    showlegend = FALSE
  )

p_risk_bar
```

::: {.obsv-box}
Observations

Under default weights, all three incidents score above the 60-point flag threshold, validating the model against the known historical cases. The file repetition rate (weight 30) and chain length (weight 25) together account for 55 points of the maximum score. Reducing either below 10 causes MellowOtter — the shortest chain — to fall below the threshold, suggesting these two features are essential for catching shorter, faster attacks.
:::

### Visualisation 10 — Weight Sensitivity Analysis

This chart shows how each incident's risk score changes as a single feature weight is varied from 0 to 50, with the remaining weight budget redistributed equally across the other four features. It reveals which features exert the most influence over whether a chain crosses the flag threshold.

```{r sensitivity-plotly, fig.height=8}
feature_names <- c("F1: Chain length", "F2: File repetition",
                   "F3: Poster reachable", "F4: Payload staged",
                   "F5: Prior cover-up")

weight_seq <- seq(0, 50, by = 5)

sensitivity_rows <- list()
for (fi in 1:5) {
  for (w in weight_seq) {
    remaining <- (100 - w) / 4
    weights   <- rep(remaining, 5)
    weights[fi] <- w

    row_scores <- risk_scores_base %>%
      mutate(
        score = weights[1] * f1_chain_length +
                weights[2] * f2_repetition +
                weights[3] * f3_poster_reachable +
                weights[4] * f4_staged +
                weights[5] * f5_coverup,
        feature        = feature_names[fi],
        feature_weight = w
      ) %>%
      select(incident, feature, feature_weight, score)

    sensitivity_rows <- append(sensitivity_rows, list(row_scores))
  }
}

sensitivity_df <- bind_rows(sensitivity_rows)

p_sensitivity <- plot_ly()

incidents_list <- c("HiddenOrca", "MellowOtter", "SwiftWren")

for (inc in incidents_list) {
  for (feat in feature_names) {
    df_sub <- sensitivity_df %>%
      filter(incident == inc, feature == feat)

    p_sensitivity <- p_sensitivity %>%
      add_trace(
        data        = df_sub,
        x           = ~feature_weight,
        y           = ~score,
        type        = "scatter",
        mode        = "lines+markers",
        name        = inc,
        legendgroup = inc,
        showlegend  = (feat == feature_names[1]),
        line        = list(color = incident_colours[inc]),
        marker      = list(color = incident_colours[inc], size = 5),
        text        = ~paste0(inc, "<br>Feature: ", feat,
                               "<br>Weight: ", feature_weight,
                               "<br>Score: ", round(score, 1)),
        hovertemplate = "%{text}<extra></extra>",
        visible     = if (feat == feature_names[1]) TRUE else "legendonly"
      )
  }
}

p_sensitivity <- p_sensitivity %>%
  layout(
    title      = list(text = "Score Sensitivity: How Each Feature Weight Affects the Final Score",
                      font = list(size = 15)),
    xaxis      = list(title = "Feature weight (0-50, others redistributed equally)"),
    yaxis      = list(title = "Risk score (0-100)", range = c(0, 105)),
    legend     = list(title = list(text = "Incident")),
    hovermode  = "closest",
    shapes     = list(list(
      type = "line", x0 = 0, x1 = 50,
      y0 = 60, y1 = 60,
      line = list(color = "#e41a1c", dash = "dash", width = 1.5)
    )),
    annotations = list(list(
      x = 48, y = 62, text = "Flag threshold (60)",
      showarrow = FALSE, font = list(color = "#e41a1c", size = 10)
    )),
    updatemenus = list(list(
      type    = "dropdown",
      active  = 0,
      buttons = lapply(seq_along(feature_names), function(i) {
        feat <- feature_names[i]
        visible_vec <- unlist(lapply(incidents_list, function(inc) {
          lapply(feature_names, function(f) f == feat)
        }))
        list(method  = "restyle",
             args    = list("visible", as.list(visible_vec)),
             label   = feat)
      }),
      x = 0, xanchor = "left", y = 1.15, yanchor = "top"
    ))
  )

p_sensitivity
```

::: {.obsv-box}
Observations

The sensitivity chart shows that F1 (chain length) and F2 (file repetition) have the steepest score gradients — small changes to their weights produce large changes in the final score, particularly for SwiftWren (the longest chain). F3 (poster reachable) has a flat gradient for HiddenOrca and MellowOtter because those chains also reached posting-capable agents, so increasing F3's weight lifts all three incidents equally. F5 (prior cover-up) has the smallest impact because it is binary and all three incidents share the same value for it.
:::

::: {.biz-box}
Implications

Organisations with a high volume of legitimate long relay chains, such as automated pipeline orchestration systems, may need to reduce the chain length weight to avoid false positives, compensating by increasing the file repetition rate weight. Organisations that have already restricted posting permissions can reduce the poster-reachable weight, as that control is already enforced at the system level. The flag threshold can also be adjusted: a lower threshold increases sensitivity (catching more attacks) at the cost of more false positives, while a higher threshold reduces noise but risks missing shorter or faster attacks like MellowOtter.
:::

---

## Conclusion

### What the Posts Mean

The three anomalous SaidIt posts have no semantic meaning as business content. Their content was pre-staged in `.txt` files by a malicious actor or compromised agent. The probable purposes are: confirming the injection mechanism works end-to-end; avoiding immediate alarm by using unrecognisable content; and maintaining non-attribution since the content cannot be linked to any legitimate business activity.

### Where the Content Came From

| Incident | Payload file | Author | Notes |
|---|---|---|---|
| HiddenOrca | `HiddenOrca.txt` | Unknown | No `create_file` event — externally injected |
| MellowOtter | `MellowOtter.txt` | `noah_mariner`'s agent | 44,879 bytes |
| SwiftWren | `SwiftWren.txt` | `emma_harbor`'s agent | 30,615 bytes |

John Windward did not author the content. His agent was the designated executor, the final node in a carefully engineered relay chain.

### Five-Point Evidence Summary

::: {.obsv-box}
Observations

Five categories of evidence support the prompt-injection conclusion:

1. Structural anomaly — `content_source` field rather than `content` in post details
2. Relay chain — hundreds of `queue_subordinate_task` events carrying the same instructions file
3. Targeting — instructions specifically named `john_windward` as terminal actor
4. Automation — 2-second trigger-to-post latency, impossible for a human
5. Cover-up — automatic deletion of both files within 5 seconds of posting
:::

---

## References

- VAST Challenge 2026 Mini-Challenge 2 problem statement and data.
- Wickham, H. et al. (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686.
- Pedersen, T.L. (2024). ggraph: An Implementation of Grammar of Graphics for Graphs and Networks. R package version 2.2.1.
- Csardi, G. & Nepusz, T. (2006). The igraph software package for complex network research. InterJournal, Complex Systems, 1695.
- Almende B.V. (2023). visNetwork: Network Visualization using vis.js Library. R package version 2.1.2.
- Sievert, C. (2020). Interactive Web-Based Data Visualization with R, plotly, and shiny. Chapman and Hall/CRC.
- Xie, Y., Cheng, J. & Tan, X. (2023). DT: A Wrapper of the JavaScript Library DataTables. R package version 0.31.
