---
title: "Hands-on Exercise 5"
author: "Huang Kuan Yu"
date: "2026-05-11"
date-modified: "last-modified" 
  
format:
  html:
    toc: true
    toc-expand: 1
    toc-title: "Contents"
    css: custom.scss
    number-sections: true 
execute:
  echo: true 
  eval: true 
  warning: false 
  freeze: true
---

## Getting Started

### Install Packages

Import all necessary packages for hands-on exercise.

```{r}
pacman::p_load(
  plotly, ggtern, corrplot, ggstatsplot,
  seriation, dendextend, heatmaply,
  GGally, parallelPlot,
  treemap, treemapify,
  tidyverse, d3treeR
)
```

## Chapter 13: Creating Ternary Plot with R

### Import Data

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv") 
```

### Prepare the data

Use the mutate() function of dplyr package to derive three new measures, namely: young, active, and old.

```{r}
#Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

### Plotting Ternary Diagram with R

### Plotting a static ternary diagram

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 12
#|fig-height: 8
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```
:::

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 12
#|fig-height: 8
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```
:::

### Plotting an interactive ternary diagram

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 12
#|fig-height: 8

label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}


axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}


axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```
:::

## Chapter 6: Visual Correlation Analysis

### Import Data

```{r}
wine <- read_csv("data/wine_quality.csv")
```

### Building correlation matrix: pairs() method

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 12
#|fig-height: 8

# Step 3: Include with correlation coefficients
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...) {
usr <- par("usr")
on.exit(par(usr = usr))
par(usr = c(0, 1, 0, 1))
r <- abs(cor(x, y, use="complete.obs"))
txt <- format(c(r, 0.123456789), digits=digits)[1]
txt <- paste(prefix, txt, sep="")
if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

# Step 1: building a basic correlation matrix 
pairs(wine[,2:12], 
# Step 2: Drawing the lower corner, can show the other half with lower.panel
      upper.panel = panel.cor)
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
pairs(wine[,1:11])
# Step 3: Include with correlation coefficients
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...) {
usr <- par("usr")
on.exit(par(usr = usr))
par(usr = c(0, 1, 0, 1))
r <- abs(cor(x, y, use="complete.obs"))
txt <- format(c(r, 0.123456789), digits=digits)[1]
txt <- paste(prefix, txt, sep="")
if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

# Step 1: building a basic correlation matrix 
pairs(wine[,2:12], 
# Step 2: Drawing the lower corner, can show the other half with lower.panel
      upper.panel = panel.cor)
```
:::

### Visualising correlation matrix: ggcormat()

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 16
#|fig-height: 8
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11,
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  title    = "Correlogram for wine dataset",
  subtitle = "Four pairs are no significant at p < 0.05"
)

ggplot.component = list(
    theme(text=element_text(size=5),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8)))
```

#### The code

```{r}
#|eval: false
#|fig-width: 16
#|fig-height: 8
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11,
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  title    = "Correlogram for wine dataset",
  subtitle = "Four pairs are no significant at p < 0.05"
)

ggplot.component = list(
    theme(text=element_text(size=5),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8)))
```
:::

### Building multiple plots

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 12
#|fig-height: 8
grouped_ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 2),
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  annotation.args = list(
    tag_levels = "a",
    title = "Correlogram for wine dataset",
    subtitle = "The measures are: alcohol, sulphates, fixed acidity, citric acid, chlorides, residual sugar, density, free sulfur dioxide and volatile acidity",
    caption = "Dataset: UCI Machine Learning Repository"
  )
)
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
grouped_ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 2),
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  annotation.args = list(
    tag_levels = "a",
    title = "Correlogram for wine dataset",
    subtitle = "The measures are: alcohol, sulphates, fixed acidity, citric acid, chlorides, residual sugar, density, free sulfur dioxide and volatile acidity",
    caption = "Dataset: UCI Machine Learning Repository"
  )
)
```
:::

### Visualising correlation matrix using corrplot package

See intruction in code

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 12
#|fig-height: 8
wine.cor <- cor(wine[, 1:11])

corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
wine.cor <- cor(wine[, 1:11])
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```
:::

### Combining corrgram with the significant test

::: panel-tabset
#### The plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
wine.num <- wine[, 1:11]

wine.cor <- cor(wine.num, use = "complete.obs")

wine.sig <- cor.mtest(wine.num, conf.level = .95)
corrplot(wine.cor,
         method = "number",
         type = "lower",
         diag = FALSE,
         tl.col = "black",
         tl.srt = 45,
         p.mat = wine.sig$p,
         sig.level = .05)
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
wine.num <- wine[, 1:11]

wine.cor <- cor(wine.num, use = "complete.obs")

wine.sig <- cor.mtest(wine.num, conf.level = .95)
corrplot(wine.cor,
         method = "number",
         type = "lower",
         diag = FALSE,
         tl.col = "black",
         tl.srt = 45,
         p.mat = wine.sig$p,
         sig.level = .05)
```
:::

### Reorder a corrgram

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 12
#|fig-height: 8
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               order="AOE",
               tl.col = "black")
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
corrplot.mixed(wine.cor, 
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               order="AOE",
               tl.col = "black")
```
:::

### Reorder using hclust

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 12
#|fig-height: 8
corrplot(wine.cor, 
         method = "ellipse", 
         tl.pos = "lt",
         tl.col = "black",
         order="hclust",
         hclust.method = "ward.D",
         addrect = 3)
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
corrplot(wine.cor, 
         method = "ellipse", 
         tl.pos = "lt",
         tl.col = "black",
         order="hclust",
         hclust.method = "ward.D",
         addrect = 3)
```
:::

## Chapter 14: Heatmap for Visualising and Analysing Multivariate Data

### Import data

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

### Prepare data

```{r}
# Convert tibble to data frame
wh_df <- as.data.frame(wh)

# Set row names using country names
row.names(wh_df) <- wh_df$Country

# Select variables
wh1 <- dplyr::select(wh_df, c(3, 7:12))

# Convert to matrix
wh_matrix <- data.matrix(wh1)
```

### Static heatmap

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 12
#|fig-height: 8
wh_heatmap <- heatmap(wh_matrix,
                      scale="column",
                      cexRow = 0.6, 
                      cexCol = 0.8,
                      margins = c(10, 4))
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
wh_heatmap <- heatmap(wh_matrix,
                      scale="column",
                      cexRow = 0.6, 
                      cexCol = 0.8,
                      margins = c(10, 4))
```
:::

### Creating interactive heatmap

::: panel-tabset
#### The plot

```{r}
#|echo: false
#|fig-width: 12
#|fig-height: 8
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

#### The code

```{r}
#|eval: false
#|fig-width: 12
#|fig-height: 8
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```
:::

### Data transformation

::: panel-tabset
### Scaling method

```{r}
#|fig-width: 12
#|fig-height: 8
heatmaply(wh_matrix[, -c(1, 2, 4, 5)],
          scale = "column")
```

### Normalising method

```{r}
#|fig-width: 12
#|fig-height: 8
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

### Percentising method

```{r}
#|fig-width: 12
#|fig-height: 8
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```
:::

### Clustering algotrithm

**heatmaply** supports a variety of hierarchical clustering algorithm. The main arguments provided are:

- *distfun*: function used to compute the distance (dissimilarity) between both rows and columns. Defaults to dist. The options “pearson”, “spearman” and “kendall” can be used to use correlation-based clustering, which uses as.dist(1 - cor(t(x))) as the distance metric (using the specified correlation method).

- *hclustfun*: function used to compute the hierarchical clustering when *Rowv* or *Colv* are not dendrograms. Defaults to *hclust*.

- *dist_method* default is NULL, which results in “euclidean” to be used. It can accept alternative character strings indicating the method to be passed to distfun. By default *distfun* is “dist”” hence this can be one of “euclidean”, “maximum”, “manhattan”, “canberra”, “binary” or “minkowski”.

- *hclust_method* default is NULL, which results in “complete” method to be used. It can accept alternative character strings indicating the method to be passed to *hclustfun*. By default hclustfun is hclust hence this can be one of “ward.D”, “ward.D2”, “single”, “complete”, “average” (= UPGMA), “mcquitty” (= WPGMA), “median” (= WPGMC) or “centroid” (= UPGMC).

In general, a clustering model can be calibrated either manually or statistically.

### Manual approach

```{r}
#| fig-width: 12
#| fig-height: 8
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "ward.D")
```

### Statistical approach

```{r}
#| fig-width: 12
#| fig-height: 8
wh_d <- dist(normalize(wh_matrix[, -c(1, 2, 4, 5)]), method = "euclidean")
dend_expend(wh_d)[[3]]
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)

heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3)
```

### Seriation

::: panel-tabset
### Optimal Leaf Ordering (OLO)

```{r}
#| fig-width: 12
#| fig-height: 8
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "OLO")
```

### Gruvaeus and Wainer

```{r}
#| fig-width: 12
#| fig-height: 8
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "GW")
```

### Mean

```{r}
#| fig-width: 12
#| fig-height: 8
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "mean")
```

### None

```{r}
#| fig-width: 12
#| fig-height: 8
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none")
```
:::

### Finishing touch:

::: panel-tabset
#### The plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          Colv=NA,
          seriate = "none",
          colors = Blues,
          k_row = 5,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="World Happiness Score and Variables by Country, 2018 \nDataTransformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries"
          )
```

#### The code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          Colv=NA,
          seriate = "none",
          colors = Blues,
          k_row = 5,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="World Happiness Score and Variables by Country, 2018 \nDataTransformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries"
          )
```
:::

## Chapter 15: Visual Multivariate Analysis with Parallel Coordinates Plot

For Chapter 15 same data as chapter 14 is used.

### Plotting Static Parallel Coordinates Plot

Things to learn from the code chunk above.

- `groupColumn` argument is used to group the observations (i.e. parallel lines) by using a single variable (i.e. Region) and colour the parallel coordinates lines by region name.

- `scale` argument is used to scale the variables in the parallel coordinate plot by using `uniminmax` method. The method univariately scale each variable so the minimum of the variable is zero and the maximum is one.

- `alphaLines` argument is used to reduce the intensity of the line colour to 0.2. The permissible value range is between 0 to 1.

- `boxplot` argument is used to turn on the boxplot by using logical `TRUE`. The default is `FALSE`.

- `title` argument is used to provide the parallel coordinates plot a title.

::: panel-tabset
### Simple parallel coordinates

```{r}
#| fig-width: 12
#| fig-height: 8
ggparcoord(data = wh, 
           columns = c(7:12))
```

### Parallel coordinates with boxplot

```{r}
#| fig-width: 12
#| fig-height: 8
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

### Parallel coordinates with facet

```{r}
#| fig-width: 12
#| fig-height: 8
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```
:::

### Rotate and adjust the x-axis label

::: panel-tabset
#### The plot

```{r}
#| echo: false 
#| fig-width: 12
#| fig-height: 8
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

#### The code

```{r}
#| echo: false 
#| fig-width: 12
#| fig-height: 8
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```
:::

### Plotting Interactive Parallel Coordinates Plot: parallelPlot methods

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
wh <- wh %>%
  select("Happiness score", c(7:12))
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             width = 320,
             height = 250,
             rotateTitle = TRUE,
             continuousCS = "YlOrRd",
             histoVisibility = histoVisibility)
```

## Chapter 16: Treemap Visualisation with R

### Import data

```{r}
realis2018 <- read_csv("data/realis2018.csv")
```

The data.frame *realis2018* is in trasaction record form, which is highly disaggregated and not appropriate to be used to plot a treemap. In this section, we will perform the following steps to manipulate and prepare a data.frtame that is appropriate for treemap visualisation:

- group transaction records by *Project Name*, *Planning Region*, *Planning Area*, *Property Type* and *Type of Sale*, and

- compute *Total Unit Sold*, *Total Area*, *Median Unit Price* and *Median Transacted Price* by applying appropriate summary statistics on *No. of Units*, *Area (sqm)*, *Unit Price (\$ psm)* and *Transacted Price (\$)* respectively.

Two key verbs of **dplyr** package, namely: *group_by()* and *summarize()* will be used to perform these steps.

*group_by()* breaks down a data.frame into specified groups of rows. When you then apply the verbs above on the resulting object they’ll be automatically applied “by group”.

Grouping affects the verbs as follows:

- grouped *select()* is the same as ungrouped *select()*, except that grouping variables are always retained.

- grouped *arrange()* is the same as ungrouped; unless you set *.by_group = TRUE*, in which case it orders first by the grouping variables.

- *mutate()* and *filter()* are most useful in conjunction with window functions (like *rank()*, or *min(x) == x*). They are described in detail in vignette(“window-functions”).

- *sample_n()* and *sample_frac()* sample the specified number/fraction of rows in each group.

- *summarise()* computes the summary for each group.

In our case, *group_by()* will used together with *summarise()* to derive the summarised data.frame.

### Data wrangling

::: panel-tabset
### Grouped summaries without the pipe

```{r}
realis2018_grouped <- group_by(realis2018, `Project Name`,
                               `Planning Region`, `Planning Area`, 
                               `Property Type`, `Type of Sale`)
realis2018_summarised <- summarise(realis2018_grouped, 
                          `Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
                          `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
                          `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE), 
                          `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

### Grouped summaries with the pipe

```{r}
realis2018_summarised <- realis2018 %>% 
  group_by(`Project Name`,`Planning Region`, 
           `Planning Area`, `Property Type`, 
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE), 
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```
:::

### Design Treemap with treemap package

### Design a static treemap

- index

  - The index vector must consist of at least two column names or else no hierarchy treemap will be plotted.

  - If multiple column names are provided, such as the code chunk above, the first name is the highest aggregation level, the second name the second highest aggregation level, and so on.

- vSize

  - The column must not contain negative values. This is because it’s vaues will be used to map the sizes of the rectangles of the treemaps.

::: panel-tabset
#### The plot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8

realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type = "value",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

#### The code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8

realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type = "value",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

### Colors in treemap package

::: panel-tabset
### The "value" type treemap

```{r}
#| fig-width: 12
#| fig-height: 8
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### The "manual"type treemap

```{r}
#| fig-width: 12
#| fig-height: 8
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

### Treemap layout

::: panel-tabset
### Working with algorithm argument

```{r}
#| fig-width: 12
#| fig-height: 8
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### Using sortID

```{r}
#| fig-width: 12
#| fig-height: 8
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "pivotSize",
        sortID = "Median Transacted Price",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

### Designing Treemap using treemapify package

### Design basic treemap

::: panel-tabset
#### The plot

```{r}
#|echo: false
#| fig-width: 12
#| fig-height: 8
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

#### The code

```{r}
#| eval: false
#| fig-width: 12
#| fig-height: 8
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

### Defining hierachy

::: panel-tabset
### Step 1: Group by planning region

```{r}
#| fig-width: 12
#| fig-height: 8
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") + 
  geom_treemap()
```

### Step 2: Group by planning area

```{r}
#| fig-width: 12
#| fig-height: 8
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap()
```

### Step 3: Adding boundary

```{r}
#| fig-width: 12
#| fig-height: 8
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```
:::

### Designing Interactive Treemap using d3treeR

### Install package

```{r}
#| eval: false 
install.packages("devtools")
install_github("timelyportfolio/d3treeR")
```

### Designing an interactive treemap

::: panel-tabset
### The static plot

```{r}
#| fig-width: 12
#| fig-height: 8
tm <- treemap(realis2018_summarised,
        index=c("Planning Region", "Planning Area"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        title="Private Residential Property Sold, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )

```

### The interactive plot code

```{r}
#| fig-width: 12
#| fig-height: 8
d3tree(tm,rootname = "Singapore" )
```
:::
