---
title: "Hands on Exercise 9"
author: "Huang Kuan Yu"
date: "2026-06-15"
date-modified: "last-modified" 
  
format:
  html:
    toc: true
    toc-expand: 1
    toc-title: "Contents"
    fig-width: 12
    fig-height: 8
    code-fold: true
    code-summary: "Show code"
    css: custom.scss
    number-sections: true

execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Getting Started

Launch necessary packages for all three chapters.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Chapter 21 Choropleth Mapping with R

### Data Import

```{r}
### Geospatial data
mpsz <- st_read(dsn = "data", 
                layer = "MP14_SUBZONE_WEB_PL")

### Attribute data
popdata <- read_csv("data/respopagesextod2011to2020.csv")
```

Examine data

```{r}
mpsz
```

### Data Wrangling

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

Join the attribute data and geospatial data together using `left_join`.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

Write the processed data into clean file.

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

### Choropleth Mapping Geospatial Data Using tmap

::: code-tip
Two approaches can be used to prepare thematic map using tmap, they are: - Plotting a thematic map quickly by using `qtm()`. - Plotting highly customisable thematic map by using tmap elements.
:::

#### Method 1: Using qtm()

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

::: obsv-box
- `tmap_mode()` with “plot” option: produce a static map
- fill argument: map the attribute (i.e. DEPENDENCY)
:::

### Method 2: Using tmap's elements

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

::: code-tip
Other arguments for `tm_borders()`: - `col` = border colour, - `lwd` = border line width. The default is 1, and - `lty` = border line type. The default is “solid”.
:::

### Data Classification Methods of tmap

#### Method 1: with built-in classification methods

::: panel-tabset
##### 5 Classes Quantile Classification

```{r}
#| code-fold: true 
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

#### equal data classification

```{r}
#| code-fold: true 
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

:::

#### Method 2: with custom break

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

### Colour Scheme

::: code-tip
Using ColourBrewer palette from the RColorBrewer package. Assign the preferred colour to *values* argument of `tm_scale_intervals()` To reverse the colour shading, add a “-” prefix.
:::

```{r}
#| code-fold: true
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

### Map Layouts

- `tm_legend()`: provided to change the placement, format and appearance of the legend
- `tmap_style()`: a wide variety of layout settings
- `tm_compass()`: used to add compass
- `tm_scale_bar()`: used to add scale bar
- `tm_grid()`: used to add grid line

```{r}
#| code-fold: true
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

### Drawing Small Multiple Choropleth Maps

#### Method 1: By assigning multiple values to at least one of the aesthetic arguments

```{r}
#| code-fold: true
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

#### Method 2: By defining a group-by variable in tm_facets()

```{r}
#| code-fold: true
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

#### Method 3: By creating multiple stand-alone maps with tmap_arrange()

```{r}
#| code-fold: true
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### Mappping Spatial Object Meeting a Selection Criterion

```{r}
#| code-fold: true
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

## Chapter 22 Visualising Geospatial Point Data

### Data Import

```{r}
sgpools <- read_csv("data/SGPools_svy21.csv")

### examine the data file

list(sgpools) 
```

### Creating a sf Data Frame from an Aspatial Data Frame

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

### Drawing Proportional Symbol Map

The code churn below will turn on the interactive mode of tmap.

```{r}
tmap_mode("view")
```

#### Starting Map Code

```{r}
#| eval: false

tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```

#### Full Map Code

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", ### variable used as the colour attribute variable
             size = "Gp1Gp2 Winnings",  ### Makes the map proportional
             col = "black",
             lwd = 1)
```

#### Faceted graphs

```{r}
#| code-fold: true
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) + 
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

## Chapter 23 Analytical Mapping

### Data Import

```{r}
NGA_wp <- read_rds("data/NGA_wp.rds")
```

### Basic Choropleth Mapping

#### Visualising distribution of non-functional water point

```{r}
#| code-fold: true

p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
             fill.scale = tm_scale_intervals(
               style = "equal",
               n = 10,
               values = "brewer.blues"),
             fill.legend = tm_legend(
               position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of functional water point by LGAs")
```

```{r}
#| code-fold: true

p2 <- tm_shape(NGA_wp) + 
  tm_polygons(fill = "total_wp", 
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1, 
             fill_alpha = 1) + 
  tm_title("Distribution of total  water point by LGAs")
```

Put two graphs together

```{r}
tmap_arrange(p2, p1, nrow = 1)

```

### Choropleth Map for Rates

Data preparation: deriving proportion of functional water points and non-functional water points

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

Plotting the map below.

```{r}
#| code-fold: true

tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) + 
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Rate map of functional water point by LGAs")
```

### Extreme Value Maps

#### Percentile map

*Data preparation:*

Step 1: Exclude records with NA. Step 2: Creating customised classification and extracting values

```{r}
#### step 1
NGA_wp <- NGA_wp %>%
  drop_na()
#### step 2
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

*Creating the get.var function*

- arguments:

  - vname: variable name (as character, in quotes)

  - df: name of sf data frame

- returns:

  - v: vector with values (without a column name)

```{r}
#| code-fold: true

percentmap <- function(vnam, df, legtitle = NA, mtitle = "Percentile Map") {
  percent <- c(0, .01, .1, .5, .9, .99, 1)
  var <- df[[vnam]]
  bperc <- quantile(var, percent, na.rm = TRUE)

  tm_shape(df) +
    tm_polygons(fill = NA) +
    tm_shape(df) +
    tm_polygons(
      fill = vnam,
      fill.scale = tm_scale_intervals(
        breaks = bperc,
        values = "brewer.blues",
        labels = c("< 1%", "1% - 10%", "10% - 50%", 
                   "50% - 90%", "90% - 99%", "> 99%")
      ),
      fill.legend = tm_legend(title = legtitle)
    ) +
    tm_borders() +
    tm_title(mtitle, position = c("right", "bottom"))
}
### Use below code chunk to run the function

percentmap("total_wp", NGA_wp)
```

#### Box map

*Creating the boxbreaks function*

- arguments:

  - v: vector with observations

  - mult: multiplier for IQR (default 1.5)

- returns:

  - bb: vector with 7 break points compute quartile and fences

```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

*Creating the get.var function*

- arguments:

  - vname: variable name (as character, in quotes)

  - df: name of sf data frame

- returns:

  - v: vector with values (without a column name)

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}

### code chunk below to test drive new function
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

*Boxmap function*

```{r}
boxmap <- function(vnam, df, 
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
          labels = c("lower outlier", 
                     "< 25%", 
                     "25% - 50%", 
                     "50% - 75%",
                     "> 75%", 
                     "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("left",
                               "top"))
}

### code chunk below to run the function
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```

